---
title: "Data Processing"
format: pdf
editor: visual
---

## IDS 702: Final Project FALL 2025

## Statistical Analysis of Election Outcomes in the USA, 2000-2024

## **Economic Data Processing**

This script reduces dimensionality of the previously downloaded raw ACS data, to make it easier to work with in the upcoming analyis.

------------------------------------------------------------------------

## **Block 1: Setup and Loading Data**.

```{r}
library(tidyverse)
library(here)

# Load the raw Census data
census_raw <- read_csv("census_economic_data_raw.csv", show_col_types = FALSE)

# Quick overview
cat("Original data dimensions:", nrow(census_raw), "rows x", ncol(census_raw), "columns\n")
cat("Year range:", min(census_raw$year), "-", max(census_raw$year), "\n")
cat("Number of states:", n_distinct(census_raw$NAME), "\n")
```

------------------------------------------------------------------------

## Block 2: Income Classification

**What this block does:**

\- Takes 14 separate income brackets

\- Creates 6 income bands

\- Each band is a percentage of total households

\- Converts raw counts into percentages (easier to compare across states)

**The 6 bands:**

\- Lower income: \<\$35k (combines 6 brackets: under \$10k through \$30-35k)

\- Lower-middle: \$35k-\$60k (combines 4 brackets)

\- Middle: \$60k-\$100k (combines 2 brackets)

\- Upper-middle: \$100k-\$150k (combines 2 brackets)

\- Upper: \$150k-\$200k (1 bracket)

\- Super-upper: \$200k+ (1 bracket)

```{r}
# Calculate income band percentages
census_income <- census_raw %>%
  mutate(
    # First, calculate total households by summing all income brackets
    total_hh = income_under_10kE + income_10k_15kE + income_15k_20kE + 
               income_20k_25kE + income_25k_30kE + income_30k_35kE +
               income_35k_40kE + income_40k_45kE + income_45k_50kE +
               income_50k_60kE + income_60k_75kE + income_75k_100kE +
               income_100k_125kE + income_125k_150kE + income_150k_200kE +
               income_200k_plusE,
    
    # Lower income: Under $35k (sum first 6 brackets)
    lower_income_hh = income_under_10kE + income_10k_15kE + income_15k_20kE + 
                      income_20k_25kE + income_25k_30kE + income_30k_35kE,
    lower_income_pct = (lower_income_hh / total_hh) * 100,
    
    # Lower-middle income: $35k-$60k (sum next 4 brackets)
    lower_middle_income_hh = income_35k_40kE + income_40k_45kE + 
                             income_45k_50kE + income_50k_60kE,
    lower_middle_income_pct = (lower_middle_income_hh / total_hh) * 100,
    
    # Middle income: $60k-$100k (sum next 2 brackets)
    middle_income_hh = income_60k_75kE + income_75k_100kE,
    middle_income_pct = (middle_income_hh / total_hh) * 100,
    
    # Upper-middle income: $100k-$150k (sum next 2 brackets)
    upper_middle_income_hh = income_100k_125kE + income_125k_150kE,
    upper_middle_income_pct = (upper_middle_income_hh / total_hh) * 100,
    
    # Upper income: $150k-$200k (single bracket)
    upper_income_hh = income_150k_200kE,
    upper_income_pct = (upper_income_hh / total_hh) * 100,
    
    # Super-upper income: $200k+ (single bracket)
    super_upper_income_hh = income_200k_plusE,
    super_upper_income_pct = (super_upper_income_hh / total_hh) * 100
  )
```

------------------------------------------------------------------------

## Block 3: Education Classification

**What this block does:**

\- Takes 7 education categories

→ Creates 3 meaningful levels

\- Groups education by major workforce divides

\- All percentages based on population 25+ years old

**The 3 levels:**

\- Non-college: High school grad + some college + associate's degree (no 4-year degree)

\- College: Bachelor's degree only

\- Graduate+: Master's + Professional degree + Doctorate (all post-bachelor's)

```{r}
# Calculate education level percentages
census_education <- census_income %>%
  mutate(
    # Non-college: HS through associate's (workers without 4-year degree)
    non_college_pop = edu_high_schoolE + edu_some_collegeE + edu_associatesE,
    non_college_pct = (non_college_pop / edu_total_25plusE) * 100,
    
    # College: Bachelor's degree only
    college_pop = edu_bachelorsE,
    college_pct = (college_pop / edu_total_25plusE) * 100,
    
    # Graduate+: All post-bachelor's degrees
    graduate_plus_pop = edu_mastersE + edu_professionalE + edu_doctorateE,
    graduate_plus_pct = (graduate_plus_pop / edu_total_25plusE) * 100,
    
    # Bonus: Combined college+ (bachelor's + graduate) for reference
    college_plus_pop = college_pop + graduate_plus_pop,
    college_plus_pct = (college_plus_pop / edu_total_25plusE) * 100
  )
```

------------------------------------------------------------------------

## Block 4: Housing Affordability

**What this block does:**

\- Creates housing cost burden metrics and homeownership rates

\- Calculates what % of renters are cost-burdened (HUD defines 30%+ as burdened)

\- Creates home value-to-income ratio (measures affordability)

**The 7 metrics:**

\- Homeownership rate: % of households that own their home

\- Rental rate: % of households that rent

\- Home value-to-income ratio: Median home price / median income (\>3.0 = affordability challenge)

\- Rent burden categories: Moderate (30-35%), High (35-50%), Severe (50%+)

```{r}
# Calculate housing affordability metrics
census_housing <- census_education %>%
  mutate(
    # Homeownership and rental rates
    homeownership_rate = (owner_occupiedE / total_householdsE) * 100,
    rental_rate = (renter_occupiedE / total_householdsE) * 100,
    
    # Rent burden categories (% of renters in each category)
    rent_burden_moderate_pct = (rent_burden_30_35pctE / rent_burden_totalE) * 100,
    rent_burden_high_pct = ((rent_burden_35_40pctE + rent_burden_40_50pctE) / 
                            rent_burden_totalE) * 100,
    rent_burden_severe_pct = (rent_burden_50pct_plusE / rent_burden_totalE) * 100,
    
    # Combined: All renters paying 35%+ of income on rent
    rent_burden_35plus_pct = ((rent_burden_35_40pctE + rent_burden_40_50pctE + 
                               rent_burden_50pct_plusE) / rent_burden_totalE) * 100,
    
    # Affordability ratio: Home price relative to income
    home_value_to_income_ratio = median_home_valueE / median_household_incomeE
  )
```

------------------------------------------------------------------------

## Block 5: Labor Market Indicators

**What this block does:**

\- Calculates employment and workforce participation rates

\- Shows what % of labor force is employed vs. unemployed

\- Measures labor force participation (how many people are working or seeking work)

**The 3 metrics:**

\- Employment rate: % of labor force that has a job

\- Unemployment rate: % of labor force actively seeking work

\- Labor force participation: % of population 16+ that is in the labor force

**Result:** 3 new labor market indicators.

```{r}
# Calculate labor market rates
census_labor <- census_housing %>%
  mutate(
    # Employment rate: % of labor force that is employed
    employment_rate = (labor_force_employedE / labor_force_totalE) * 100,
    
    # Unemployment rate: % of labor force that is unemployed
    unemployment_rate = (labor_force_unemployedE / labor_force_totalE) * 100,
    
    # Labor force participation: % of civilian pop 16+ in labor force
    labor_force_participation = (labor_force_totalE / civilian_pop_16plusE) * 100
  )
```

------------------------------------------------------------------------

## Block 6: Poverty and Economic Distress

**What this block does:**

\- Calculates poverty rate (% below federal poverty line)

\- Creates economic distress score combining poverty rate AND median income

\- Identifies areas with concentrated economic hardship

**The classification logic:**

\- High distress: Poverty rate \>15% AND median income \<\$45k (both conditions)

\- Moderate distress: Poverty rate \>12% OR median income \<\$50k (either condition)

\- Low distress: Neither condition met

```{r}
# Calculate poverty rate and economic distress
census_poverty <- census_labor %>%
  mutate(
    # Poverty rate: % of population below federal poverty line
    poverty_rate = (poverty_below_lineE / poverty_total_popE) * 100,
    
    # Economic distress score: categorical variable
    economic_distress_score = case_when(
      poverty_rate > 15 & median_household_incomeE < 45000 ~ "High",
      poverty_rate > 12 | median_household_incomeE < 50000 ~ "Moderate",
      TRUE ~ "Low"
    )
  )
```

------------------------------------------------------------------------

## Block 7: Demographics

**What this block does:**

\- Calculates age structure (youth, working-age, senior populations)

\- Measures geographic mobility (who moved in the past year)

\- Shows population stability vs. migration patterns

**The 7 metrics:**

\- Age groups: % under 18, % 18-64, % 65+

\- Mobility: % moved within state, % moved from another state, % moved from abroad

\- Total mobility rate: % of population that moved (all types combined)

```{r}
# Calculate demographic percentages
census_final <- census_poverty %>%
  mutate(
    # Age structure percentages
    youth_pct = (age_under_18E / total_populationE) * 100,
    working_age_pct = (age_18_64E / total_populationE) * 100,
    senior_pct = (age_65_plusE / total_populationE) * 100,
    
    # Mobility indicators (% who moved in past year)
    moved_within_state_pct = (moved_within_stateE / total_pop_1yr_plusE) * 100,
    moved_from_other_state_pct = (moved_from_other_stateE / total_pop_1yr_plusE) * 100,
    moved_from_abroad_pct = (moved_from_abroadE / total_pop_1yr_plusE) * 100,
    
    # Total mobility: all movers combined
    total_mobility_rate = ((moved_within_stateE + moved_from_other_stateE + 
                           moved_from_abroadE) / total_pop_1yr_plusE) * 100
  )
```

------------------------------------------------------------------------

## Block 8: Create Clean Processed Dataset

**What this block does:**

\- Selects only the 38 variables we need for analysis

\- Keeps identifiers, key original measures, and all new categorical variables

\- Drops the detailed raw counts that were aggregated

\- Sorts by year and state for clean organization

**What gets kept:**

\- 3 identifiers (year, GEOID, NAME)

\- 4 original measures (population, median income, per capita income, median age)

\- 2 original housing costs (median home value, median rent)

\- 28 NEW categorical variables we just created

```{r}
# Select final variables for processed dataset
census_processed <- census_final %>%
  select(
    # Identifiers
    year, GEOID, NAME,
    
    # Population
    total_populationE,
    
    # Income measures
    median_household_incomeE, per_capita_incomeE,
    lower_income_pct, lower_middle_income_pct, middle_income_pct,
    upper_middle_income_pct, upper_income_pct, super_upper_income_pct,
    
    # Education
    non_college_pct, college_pct, graduate_plus_pct, college_plus_pct,
    
    # Housing
    homeownership_rate, rental_rate,
    median_home_valueE, median_gross_rentE,
    home_value_to_income_ratio,
    rent_burden_moderate_pct, rent_burden_high_pct, rent_burden_severe_pct,
    rent_burden_35plus_pct,
    
    # Labor market
    employment_rate, unemployment_rate, labor_force_participation,
    
    # Poverty
    poverty_rate, economic_distress_score,
    
    # Demographics
    median_ageE, youth_pct, working_age_pct, senior_pct,
    total_mobility_rate, moved_within_state_pct,
    moved_from_other_state_pct, moved_from_abroad_pct
  ) %>%
  arrange(year, NAME)

# Show dimensions
cat("Final processed dataset:", nrow(census_processed), "rows ×", 
    ncol(census_processed), "columns\n")
```

------------------------------------------------------------------------

## Block 9: Data Quality Check

**What this block does:**

\- Checks for missing values across all variables

\- Verifies that income percentages sum to \~100% (validates our calculations)

\- Shows summary statistics for key economic indicators

**Why this matters:**

\- Ensures our calculations are correct

\- Identifies any data quality issues

\- Confirms percentages are in valid ranges

```{r}
# Check for missing values
cat("=== MISSING VALUES CHECK ===\n\n")
missing_summary <- census_processed %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_count))

if (nrow(missing_summary) > 0) {
  print(missing_summary)
} else {
  cat("✓ No missing values detected\n")
}

# Verify income percentages sum to 100%
cat("\n=== INCOME BAND VALIDATION ===\n")
income_check <- census_processed %>%
  mutate(
    income_total = lower_income_pct + lower_middle_income_pct + 
                   middle_income_pct + upper_middle_income_pct + 
                   upper_income_pct + super_upper_income_pct
  ) %>%
  summarise(
    min = min(income_total, na.rm = TRUE),
    mean = mean(income_total, na.rm = TRUE),
    max = max(income_total, na.rm = TRUE)
  )
print(income_check)
cat("✓ Should be ~100% (allows for rounding)\n")

# Summary statistics
cat("\n=== KEY INDICATORS SUMMARY ===\n")
census_processed %>%
  select(median_household_incomeE, poverty_rate, college_plus_pct, 
         homeownership_rate, unemployment_rate) %>%
  summary()
```

------------------------------------------------------------------------

## Block 10: Save Processed Data

**What this block does:** - Saves the final processed dataset as a CSV file - Makes it available for your economic analysis

**Output:** - `census_economic_data_processed.csv` - Your clean, analysis-ready dataset

**Result:** CSV file with 624 rows × 38 columns ready to use.

```{r}
# Save processed dataset
write_csv(census_processed, "census_economic_data_processed.csv")

cat("✓ Saved: census_economic_data_processed.csv\n")
cat("\nProcessed dataset ready for analysis!\n")
cat("  - 624 observations (52 states × 12 years)\n")
cat("  - 38 variables (down from 107)\n")
cat("  - 28 new categorical variables created\n")
```

------------------------------------------------------------------------
