---
title: "Economic Downturn, Turnout, and Incumbent Support"
subtitle: "Final Report – IDS 702 Statistical Analysis Project"
author: "Diwas Purin, Tony Ngari, Tea Tafaj, Xinhao Zhang"
date: today
format: 
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    include-in-header:
      text: |
        \usepackage{hanging}
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5,
  dpi = 300
)

library(tidyverse)
library(lfe)
library(fixest)
library(stargazer)
library(knitr)
library(gridExtra)

theme_set(theme_minimal(base_size = 12))

# load merged panel
df <- read_csv("../data/processed/merged_voting_unemployment_complete.csv", show_col_types = FALSE)

# --- RQ1 Data Prep ---
df_rq1 <- df %>%
  filter(
    VOTER_TURNOUT_PCT > 0,
    !is.na(VOTER_TURNOUT_PCT),
    !is.na(unemployment_rate),
    VOTER_TURNOUT_PCT <= 1.0
  ) %>%
  mutate(
    # Election type
    is_presidential = ifelse(YEAR %% 4 == 0, 1, 0),
    election_type   = ifelse(is_presidential == 1, "Presidential", "Midterm"),
    
    # Unemployment tiers
    unemp_mean   = mean(unemployment_rate, na.rm = TRUE),
    unemp_sd     = sd(unemployment_rate,  na.rm = TRUE),
    unemp_zscore = (unemployment_rate - unemp_mean) / unemp_sd,
    
    unemployment_tier = case_when(
      unemp_zscore < -1.0 ~ "Very Low (< -1 SD)",
      unemp_zscore < -0.5 ~ "Low (-1 to -0.5 SD)",
      unemp_zscore <  0.5 ~ "Medium (-0.5 to 0.5 SD)",
      unemp_zscore <  1.0 ~ "High (0.5 to 1 SD)",
      TRUE                ~ "Very High (> 1 SD)"
    ),
    unemployment_tier = factor(
      unemployment_tier,
      levels = c("Very Low (< -1 SD)", "Low (-1 to -0.5 SD)", "Medium (-0.5 to 0.5 SD)", "High (0.5 to 1 SD)", "Very High (> 1 SD)"),
      ordered = TRUE
    ),
    
    # Standardized unemployment
    unemployment_std = (unemployment_rate - mean(unemployment_rate)) / sd(unemployment_rate),
    
    # County type
    county_type = case_when(
      CVAP > 50000 ~ "Urban",
      CVAP > 10000 ~ "Suburban",
      TRUE         ~ "Rural"
    )
  )

# --- RQ2 Data Prep ---
incumbent_table <- tibble(
  YEAR = c(2000, 2004, 2008, 2012, 2016, 2020),
  incumbent_party = c("DEM", "REP", "REP", "DEM", "DEM", "REP")
)

df_rq2 <- df %>%
  left_join(incumbent_table, by = "YEAR") %>%
  mutate(
    winner_party = case_when(
      PRES_DEM_VOTES > PRES_REP_VOTES ~ "DEM",
      PRES_REP_VOTES > PRES_DEM_VOTES ~ "REP",
      TRUE                            ~ NA_character_
    ),
    incumbent_win = if_else(winner_party == incumbent_party, 1, 0)
  ) %>%
  filter(
    YEAR %in% c(2000, 2004, 2008, 2012, 2016, 2020),
    !is.na(incumbent_win),
    !is.na(unemployment_rate)
  ) %>%
  mutate(
    urbanicity = case_when(
      FIPS %% 1000 < 200 ~ "Urban",
      FIPS %% 1000 < 500 ~ "Suburban",
      TRUE               ~ "Rural")
  )
```

# Executive Summary

Local economic conditions influence both *who votes* and *which party voters support*, but in very different ways. Using county-level data from 2000–2022, we examine how changes in unemployment relate to **voter turnout** and **support for the incumbent presidential party**.

\
Across election cycles, turnout is shaped primarily by election type and stable county characteristics. Local economic downturns have only a weak and modest effect: when unemployment rises within a county, turnout increases slightly, but the change is small. Rural, suburban, and urban counties show different patterns, yet the practical impact of unemployment on participation remains limited.

\
In contrast, deteriorating local economic conditions meaningfully reduce support for the incumbent presidential party. Counties experiencing worse-than-usual unemployment are less likely to vote for the president’s party, even after accounting for long-standing political differences and national election dynamics. This economic penalty is strongest in rural counties and most muted in suburban and urban areas.\
Recent elections, however, reveal exceptions—some economically distressed rural counties continued supporting Republican incumbents—reflecting broader partisan realignment.\
\
**Overall Insight:** Economic downturns do little to alter *who turns out*, but they substantially shape *how counties vote*. The clearest pattern is retrospective accountability: when local economic conditions worsen, incumbents generally face electoral punishment.

\newpage

# Stakeholder

The primary stakeholders for this memo are campaign strategists and political analysts.

The objective is to enhance their understanding of the following three items.

1.  Whether local economic downturns mobilize new voters to the polls (Turnout).
2.  Whether voters consistently punish incumbents for local economic pain (Vote Choice).
3.  How these patterns differ across Rural, Suburban, and Urban battlegrounds to optimize resource allocation.

The memo provides high-level insights to guide campaign **resource allocation**, **geographic targeting**, and **message strategy**.\
All technical specifications, modeling details, and diagnostics are deferred to the Technical Appendix.

# Background and Motivation

Political theory suggests that populations differ in their typical level of political engagement. Almond and Verba’s classic “Civic Culture” framework describes a participation pyramid in which the largest group consists of disengaged citizens (“parochials”), followed by moderately involved voters (“subjects”), and finally a small group of highly engaged participants. In many systems, this bottom tier — often described as a “silent majority” — remains politically inactive unless economic or social conditions deteriorate. Under worsening economic conditions, previously uninvolved citizens may become mobilized to vote, either to punish the incumbent or express dissatisfaction with the status quo. This idea motivates our investigation of whether local economic downturns, measured through rising unemployment, affect political participation and incumbent electoral performance at the county level.

# Data Description

Our analysis uses a county-level dataset combining data from the **National Neighborhood Data Archive (NaNDA)** (Clary et al., 2024) and the **Bureau of Labor Statistics Local Area Unemployment Statistics (LAUS)** (BLS, 2024):

-   **General election turnout** for all U.S. counties from 2004–2022.
-   **Presidential election results**, used to determine whether the incumbent party carried each county.
-   **Local economic conditions**, measured through county-year unemployment rates.
-   **County classifications** (urban, suburban, rural), constructed using population and voting-eligible population indicators to assess heterogeneous responses.

The dataset spans more than 20 years of election cycles and includes both presidential and midterm elections.\
Details on data construction, variable definitions, merging procedures, and coding steps appear in the Technical Appendix.

# Research Question 1 Key Findings

**RQ1:** *How does local unemployment relate to voter turnout in U.S. elections at the county level?*

-   **Interaction Effects are Significant:** The relationship between unemployment and voter turnout differs meaningfully across rural, suburban, and urban counties, confirming that local context shapes how communities respond to economic stress.

-   **Modest Substantive Impact:** Even under recession-like conditions (a 2% increase in unemployment), predicted turnout shifts remain below 0.5 percentage points across all county types.

-   **Rural Baseline:** In rural counties (reference group), a 1% rise in unemployment is linked to a negligible increase in turnout (\~0.04 percentage points), reflecting the stability of participation in these areas.

-   **Urban and Suburban Response:** Urban and suburban counties show modestly larger positive responses, though still small in magnitude.

-   **Overall Interpretation:** Offset mobilizing and discouraging forces likely produce near-zero net effects on turnout. Local economic downturns appear to influence how voters choose rather than whether they participate, a pattern explored further in Research Question 2.

# Research Question 2 Key Findings

**RQ2:** *Does local economic downturn affect the probability that a county supports the incumbent presidential party?*

-   **Local economic downturns reduce incumbent support:** When unemployment in a county is higher than usual for that county, the incumbent is noticeably less likely to carry it, even after accounting for long-standing political differences and overall national trends.

-   **Effects are concentrated under severe distress:** At low and moderate unemployment levels, incumbent support changes little, but in counties with high and especially very high unemployment, support for the incumbent drops sharply.

-   **Rural counties are most responsive:** Across the unemployment distribution, rural areas show the largest declines in incumbent support as unemployment rises, while the relationship is more muted in suburban and urban counties.

-   **Partisan realignment creates exceptions:** In 2016 and 2020, some high-unemployment rural counties actually showed **higher** support for the Republican incumbent, reflecting a broader partisan realignment in which party loyalty can override local economic pain.

-   **Overall interpretation:** The pattern is broadly consistent with retrospective economic voting, but accountability weakens where partisan loyalty dominates economic evaluations.

# Policy Implications

-   Economic conditions play a **limited** role in shaping voter turnout but have a clearer impact on support for the incumbent presidential party, especially when unemployment rises.

-   Campaigns should target **economic messaging** toward counties experiencing severe distress, with particular attention to **rural areas**, which are most responsive to worsening conditions.

-   Electoral strategies should also recognize that urban, suburban, and rural counties **react differently** to economic stress, and places undergoing partisan realignment may require alternative communication approaches.

-   Mobilization efforts will remain more effective when focused on **structural and institutional factors**, such as registration and voting access, rather than relying solely on economic improvements to boost participation.

# Limitations

-   **Aggregate data:** The analysis relies on county-level data, which cannot capture individual-level voting behavior or underlying psychological mechanisms.

-   **Limited economic measures:** Unemployment captures only one dimension of local economic conditions. Other factors—such as wages, inflation, or industry decline—may also shape voter behavior but are not included.

-   **Timing and salience:** Annual unemployment averages may not fully reflect voters’ most salient economic experiences, especially when shocks occur close to an election.

-   **Partisan realignment:** While county fixed effects control for stable political differences, they cannot fully account for long-term partisan shifts, particularly the rural–urban realignment after 2012.

-   **Causal interpretation:** Despite a fixed-effects design, the results remain associational; unobserved local shocks may influence both economic conditions and electoral outcomes.

All methodological limitations, robustness checks, and model diagnostics are documented in the Technical Appendix.

\newpage

# Appendix

## 1. Data Sources and Construction

We construct a county–year panel dataset by merging turnout, presidential election results, and local unemployment data.

-   **Turnout data:** County-level general election turnout for U.S. counties (2004–2022).
-   **Presidential election results:** Democratic and Republican vote totals, used to construct `winner_party` and `incumbent_party`.
-   **Local economic conditions:** County unemployment rates from the Bureau of Labor Statistics (BLS).
-   **County classifications:** Urban / Suburban / Rural categories based on county population / voting-eligible population.
-   **Time period:**
    -   **Turnout analysis (RQ1):** 2004–2022
    -   **Incumbent support (RQ2):** Presidential elections in 2000, 2004, 2008, 2012, 2016, 2020

## 2. Variable Definitions

-   **Voter Turnout Rate:** The proportion of the citizen voting-age population (CVAP) that cast a ballot.

-   **Unemployment Rate:** The percentage of the labor force that is unemployed (annual average).

-   **Unemployment Tier:** Statistical classification of unemployment severity (low / medium / high) based on historical deviations.

-   **Election Type:** Indicator for presidential vs. midterm elections.

-   **County Type:** Classification as Urban, Suburban, or Rural based on voting-age population size.

-   **Incumbent Win:** Indicator (1/0) if the county voted for the incumbent presidential party.

-   **Winner Party:** The party (Democrat/Republican) that received the most votes in the county.

-   **Incumbent Party:** The party holding the presidency at the time of the election.

-   `FIPS`: County identifier for fixed effects.

## 3. Cleaning and Processing Workflow

1.  Import data and harmonize county-level identifiers.
2.  Merge turnout with unemployment data.
3.  Merge presidential election results with unemployment.
4.  Create derived variables (tiers, incumbent indicators, county type).
5.  Filter to analysis subsets:
    -   All general elections for RQ1
    -   Only presidential elections for RQ2
6.  Handle missing/unreliable entries (NA removal, turnout = 0).
7.  Convert categorical variables to factors required for FE models.

## 4. Modeling Procedures

### **RQ1: Turnout and Local Unemployment**

**Primary Model: Fixed Effects with Interaction**

We estimate a **Two-Way Fixed Effects (TWFE)** model to isolate the within-county effect of unemployment on turnout, controlling for time-invariant county characteristics and national election trends.

$$ \text{Turnout}_{it} = \beta_1 \text{Unemployment}_{it} + \beta_2 (\text{Unemployment}_{it} \times \text{CountyType}_i) + \alpha_i + \delta_t + \epsilon_{it} $$

-   $\alpha_i$ (County Fixed Effects): Controls for stable characteristics like demographics, political culture, and geography.
-   $\delta_t$ (Year Fixed Effects): Controls for national waves (e.g., 2008 surge, 2014 slump).
-   **Interaction Term:** Allows the effect of unemployment to vary by Rural (Reference), Suburban, and Urban classifications.
-   **F-Test:** Conducted to formally test if the interaction terms significantly improve model fit compared to a base model.

### **RQ2: Economic Downturn and Incumbent Support**

**Fixed-effects logistic regression:**

We focus only on **presidential election years** (2000–2020). For each county and election, we identify:

-   which party won the county (Democrat or Republican),

-   which party held the presidency in that year, and

-   whether the county voted **for** the incumbent party.

We examine how changes in local unemployment relate to the likelihood that a county supports the incumbent presidential party. To do this fairly, we compare each county **to itself over time**, rather than comparing counties to one another. This helps account for stable differences across places—such as long-term political leanings or demographic makeup—that could otherwise bias the results.

We estimate a model that looks at how support for the incumbent changes in years when unemployment in a county is higher or lower than usual. We also account for national factors—such as political climate, incumbent popularity, or major economic events—that affect all counties in the same election year.

In simple terms, this model tells us whether counties tend to **punish or reward** the incumbent party when their own local economic conditions worsen.

## 5. Regression Tables

::: {style="page-break-after: always;"}
:::

### Table 1: Effect of Unemployment on Voter Turnout (RQ1)

```{r regression-table-rq1, results='asis'}
# Load the saved model objects from the analysis pipeline
load("../data/output/regression_models.RData")

# Determine output format for stargazer
table_type <- ifelse(knitr::is_latex_output(), "latex", "html")

stargazer(model_full,
          type = table_type,
          title = "Effect of Unemployment on Voter Turnout (Fixed Effects Model)",
          model.numbers = FALSE,
          dep.var.labels = "Voter Turnout Rate",
          covariate.labels = c("Unemployment Rate (Rural baseline)", 
                               "Unemp × Suburban", 
                               "Unemp × Urban"),
          omit.stat = c("ser"),
          digits = 4,
          single.row = FALSE,
          keep.stat = c("n", "rsq", "adj.rsq"),
          notes = c("Standard errors in parentheses.",
                    "Model includes county and year fixed effects.",
                    "Reference category: Rural counties."),
          notes.append = TRUE,
          header = FALSE)
```

### **Table 2: Effect of Unemployment on Incumbent Support (RQ2)**

```{r}
library(kableExtra)

# Fit model
model_rq2 <- fixest::feglm(
  incumbent_win ~ unemployment_rate | FIPS + YEAR,
  family = "binomial",
  data = df_rq2,
  vcov = ~ FIPS
)

# Extract components
beta <- coef(model_rq2)["unemployment_rate"]
se   <- sqrt(diag(vcov(model_rq2)))["unemployment_rate"]
z    <- beta / se
p    <- 2 * pnorm(abs(z), lower.tail = FALSE)
odds <- exp(beta)

coef_tbl <- data.frame(
  Term        = "Unemployment Rate",
  Estimate    = sprintf("%.3f", beta),
  Std.Error   = sprintf("%.3f", se),
  z.value     = sprintf("%.3f", z),
  p.value     = sprintf("%.3f", p),
  Odds.Ratio  = sprintf("%.3f", odds)
)

kable(
  coef_tbl,
  caption = "Effect of Local Unemployment on Support for the Incumbent Party (Fixed-Effects Logistic Model)",
  booktabs = TRUE,
  align = "lccccc"
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE
  )


```

This model includes county fixed effects and election-year fixed effects, which control for differences across counties and for national conditions that affect all counties in the same year. Because these factors are controlled for in the background, unemployment is the only predictor the model estimates directly and the only variable that appears in the table. **Interpretation:** A one-percentage-point increase in unemployment reduces the odds that the incumbent party wins the county by about 9%, holding county characteristics and national election conditions constant.

## 6. Supplementary Figures

### Figure S1: Turnout Trends by Election Type

```{r}
ggplot(df_rq1, aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT, 
               color = election_type)) +
  geom_point(alpha = 0.1, size = 1) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_manual(values = c("Presidential" = "#2171b5", "Midterm" = "#cb181d")) +
  labs(
    title = "Unemployment vs Turnout by Election Type",
    subtitle = "Analysis Period: 2004-2022",
    x = "Unemployment Rate (%)",
    y = "Voter Turnout Rate",
    color = "Election Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The graph shows that Turnout is driven more by election cycle than economic conditions.

### Figure S2: Distribution of Turnout by Economic Condition

```{r fig.height=3.5, fig.width=7}
ggplot(df_rq1, aes(x = county_type, y = VOTER_TURNOUT_PCT, fill = unemployment_tier)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.5) +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.8, name = "Unemployment Tier") +
  labs(
    title = "Voter Turnout Distribution by County Type",
    subtitle = "Comparing turnout across different levels of economic stress",
    x = "County Type",
    y = "Voter Turnout Rate"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

The graph indicates that Turnout distributions are stable across economic tiers.

### Figure S3: Unemployment vs Turnout by County Type

```{r fig.height=3.5, fig.width=7}
ggplot(df_rq1, aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT, 
               color = county_type)) +
  geom_point(alpha = 0.1, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Unemployment vs Turnout by County Type",
    subtitle = "Interaction: Unemployment Rate × Urban/Rural Classification",
    x = "Unemployment Rate (%)",
    y = "Voter Turnout Rate",
    color = "County Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

This figure indicates the interaction effect: the relationship between unemployment and turnout varies by county type.

### **Figure S4: Incumbent support vs local unemployment**

```{r fig.height=3.5, fig.width=7}
# Figure S4: Incumbent Support by Local Unemployment (Weighted LOESS)

ggplot(df_rq2, aes(x = unemployment_rate, y = incumbent_win)) +
  geom_jitter(alpha = 0.1, width = 0.15, height = 0.02) +
  geom_smooth(
    method = "loess",
    se = TRUE,
    aes(weight = CVAP),
    size = 1,
    color = "#1f78b4"
  ) +
  coord_cartesian(xlim = c(0, 15)) +
  scale_y_continuous(limits = c(0, 1),
                     breaks = seq(0, 1, 0.2),
                     name = "Share voting for incumbent (0–1)") +
  scale_x_continuous(name = "Local unemployment rate (%)") +
  labs(
    title = "Incumbent Support by Local Unemployment",
    subtitle = "Population-weighted outcomes with smoothed trend (trimmed to realistic unemployment range)"
  ) +
  theme_minimal(base_size = 12)



```

This figure shows that counties with higher unemployment are generally less likely to support the incumbent presidential party, with the smooth trend line highlighting the overall downward relationship once extreme values are trimmed.

### **Figure S5: Incumbent support vs unemployment, by county type**

```{r fig.height=3.5, fig.width=7}
# Figure S5: Incumbent Support, Unemployment, and County Type (Weighted LOESS)

ggplot(df_rq2, aes(x = unemployment_rate, y = incumbent_win)) +
  geom_jitter(alpha = 0.10, width = 0.15, height = 0.02) +
  geom_smooth(
    method = "loess",
    se = TRUE,
    aes(weight = CVAP),
    size = 1,
    color = "#1f78b4"
  ) +
  facet_wrap(~ urbanicity, nrow = 1) +
  coord_cartesian(xlim = c(0, 15)) +
  scale_y_continuous(limits = c(0, 1),
                     breaks = seq(0, 1, 0.2),
                     name = "Share voting for incumbent (0–1)") +
  scale_x_continuous(name = "Local unemployment rate (%)") +
  labs(
    title = "Incumbent Support, Unemployment, and County Type",
    subtitle = "Population-weighted LOESS trends across county types (trimmed to realistic unemployment range)"
  ) +
  theme_minimal(base_size = 12)


```

Across rural, suburban, and urban areas, higher unemployment is consistently linked to lower incumbent support, though the strength and shape of the relationship differ across county types, reflecting the distinct political and economic contexts of each.

\newpage

# References

\begin{hangparas}{0.5in}{1}

Bureau of Labor Statistics. (2024). \textit{Local Area Unemployment Statistics (LAUS)}. U.S. Department of Labor. https://www.bls.gov/lau/

Clary, W., Gomez-Lopez, I. N., Chenoweth, M., Gypin, L., Clarke, P., Noppert, G., Li, M., \& Kollman, K. (2024). \textit{National Neighborhood Data Archive (NaNDA): Voter Registration, Turnout, and Partisanship by County, United States, 2004-2022} (ICPSR 38506, Version 2). Inter-university Consortium for Political and Social Research. https://doi.org/10.3886/ICPSR38506.v2

Fiorina, M. P. (1981). \textit{Retrospective Voting in American National Elections}. Yale University Press.

Rosenstone, S. J. (1982). Economic adversity and voter turnout. \textit{American Journal of Political Science}, \textit{26}(1), 25--46. https://doi.org/10.2307/2110837

Schlozman, K. L., \& Verba, S. (1981). \textit{Injury to Insult: Unemployment, Class, and Political Response}. Harvard University Press.

Almond, Gabriel A., and Sidney Verba. The Civic Culture: Political Attitudes and Democracy in Five Nations. Princeton University Pre

Berelson, Bernard R., Paul F. Lazarsfeld, and William N. McPhee. Voting: A Study of Opinion Formation in a Presidential Campaign. University of Chicago Press, 1954
\end{hangparas}

---
title: "Unemployment and Voter Turnout Analysis"
subtitle: "Research Question 1: How does local economic downturn relate to voter turnout?"
author: "Statistical Analysis - STA 702"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    self-contained: true
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# **Theoretical Motivation - Political Participation Framework**

Political theory suggests that populations differ in their typical level of political engagement. Almond and Verba’s classic “Civic Culture” framework describes a participation pyramid in which the largest group consists of disengaged citizens (“parochials”), followed by moderately involved voters (“subjects”), and finally a small group of highly engaged participants. In many systems, this bottom tier — often described as a “silent majority” — remains politically inactive unless economic or social conditions deteriorate. Under worsening economic conditions, previously uninvolved citizens may become mobilized to vote, either to punish the incumbent or express dissatisfaction with the status quo. This idea motivates our investigation of whether local economic downturns, measured through rising unemployment, affect the incumbent party’s electoral performance at the county level.\
\
Almond, Gabriel A., and Sidney Verba. *The Civic Culture: Political Attitudes and Democracy in Five Nations.* Princeton University Press, 1963

Berelson, Bernard R., Paul F. Lazarsfeld, and William N. McPhee. *Voting: A Study of Opinion Formation in a Presidential Campaign.* University of Chicago Press, 1954.

```{r}
library(ggplot2)
library(dplyr)

# Data: three layers of the participation pyramid
pyramid_df <- tibble(
  level = factor(c("Participants", "Subjects", "Parochials"), 
                 levels = c("Participants", "Subjects", "Parochials")),
  ymin = c(0.66, 0.33, 0),
  ymax = c(1, 0.66, 0.33),
  label = c(
    "Participants:\nHighly engaged voters;\nconsistent political activity",
    "Subjects:\nModerately engaged;\nresponsive in major elections",
    "Parochials:\nLow engagement;\n'Silent Majority'\nactivated by crises"
  )
)

# Base triangle shape
triangle <- data.frame(
  x = c(0.5, 1, 0),
  y = c(1, 0, 0)
)

ggplot() +
  # Triangle outline
  geom_polygon(data = triangle, aes(x, y), fill = "white", color = "black") +
  
  # Layers inside the triangle
  geom_rect(data = pyramid_df, 
            aes(xmin = 0, xmax = 1, ymin = ymin, ymax = ymax, fill = level),
            alpha = 0.6, color = "black") +
  
  # Descriptive text inside each layer
  geom_text(data = pyramid_df,
            aes(x = 0.5, y = (ymin + ymax) / 2, label = label),
            color = "black", size = 4, lineheight = 1,
            fontface = "bold") +
  
  scale_fill_manual(values = c(
    "Participants" = "#6fa8dc",
    "Subjects" = "#9fc5e8",
    "Parochials" = "#cfe2f3"
  )) +
  
  coord_fixed() +
  theme_void() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  ) +
  ggtitle("Political Participation Pyramid")





```

# Introduction

This analysis investigates the relationship between local economic conditions (measured by unemployment rate) and voter turnout in U.S. presidential and midterm elections from 2004-2022.

**Research Question:** How does local economic downturn relate to voter turnout in U.S. elections at the county level?

**Key Variables:**

-   **Outcome:** `VOTER_TURNOUT_PCT` = Voter turnout rate (continuous)
-   **Predictor:** `unemployment_rate` = County unemployment rate (%)
-   **Interaction:** `unemployment_rate × county_type` (Urban/Rural classification)

# Data Preparation

```{r load-packages}
# Load required packages
library(tidyverse)
library(lfe)           # For fixed effects regression
library(stargazer)     # For regression tables
library(ggplot2)
library(knitr)
library(gridExtra)

# Set ggplot theme
theme_set(theme_minimal(base_size = 12))
```

```{r load-data}
# Load the merged dataset
df <- read_csv("../data/processed/merged_voting_unemployment_complete.csv",
               show_col_types = FALSE)

# Display data structure
cat("Dataset Overview:\n")
cat("Observations:", nrow(df), "\n")
cat("Counties:", length(unique(df$FIPS)), "\n")
cat("Years:", min(df$YEAR), "-", max(df$YEAR), "\n")
cat("Variables:", ncol(df), "\n")
```

## Variable Creation

```{r create-variables}
df <- df %>%
  mutate(
    # Election type indicator
    is_presidential = ifelse(YEAR %% 4 == 0, 1, 0),
    election_type = ifelse(is_presidential == 1, "Presidential", "Midterm"),
    
    # Economic downturn indicator (unemployment > median)
    median_unemp = median(unemployment_rate, na.rm = TRUE),
    economic_downturn = ifelse(unemployment_rate > median_unemp, 1, 0),
    downturn_label = ifelse(economic_downturn == 1, 
                           "High Unemployment", "Low Unemployment"),
    
    # Standardized variables
    unemployment_std = (unemployment_rate - mean(unemployment_rate)) / 
                       sd(unemployment_rate),
    
    # Urban/rural classification based on CVAP
    # Large counties (>50k CVAP) = Urban
    # Medium counties (10k-50k) = Suburban  
    # Small counties (<10k) = Rural
    county_type = case_when(
      CVAP > 50000 ~ "Urban",
      CVAP > 10000 ~ "Suburban",
      TRUE ~ "Rural"
    ),
    
    # Factor variables for regression
    FIPS_factor = as.factor(FIPS),
    YEAR_factor = as.factor(YEAR),
    STATE_factor = as.factor(STATE_FIPS)
  )

cat("✓ Created classification variables\n")
cat("  - Median unemployment:", round(df$median_unemp[1], 2), "%\n")
```

# Descriptive Statistics

## Summary Statistics

```{r summary-stats}
# Overall summary
df %>%
  select(VOTER_TURNOUT_PCT, unemployment_rate, CVAP) %>%
  summary() %>%
  kable(caption = "Summary Statistics: Key Variables", digits = 3)
```

## Distribution by Groups

```{r stats-by-group}
# By election type
df %>%
  group_by(election_type) %>%
  summarise(
    N = n(),
    Avg_Turnout = mean(VOTER_TURNOUT_PCT, na.rm = TRUE),
    SD_Turnout = sd(VOTER_TURNOUT_PCT, na.rm = TRUE),
    Avg_Unemployment = mean(unemployment_rate, na.rm = TRUE),
    SD_Unemployment = sd(unemployment_rate, na.rm = TRUE)
  ) %>%
  kable(caption = "Statistics by Election Type", digits = 3)

# By county type
df %>%
  group_by(county_type) %>%
  summarise(
    N = n(),
    Counties = n_distinct(FIPS),
    Avg_Turnout = mean(VOTER_TURNOUT_PCT, na.rm = TRUE),
    SD_Turnout = sd(VOTER_TURNOUT_PCT, na.rm = TRUE),
    Avg_Unemployment = mean(unemployment_rate, na.rm = TRUE),
    SD_Unemployment = sd(unemployment_rate, na.rm = TRUE)
  ) %>%
  kable(caption = "Statistics by County Type", digits = 3)

# By economic downturn status
df %>%
  group_by(downturn_label) %>%
  summarise(
    N = n(),
    Avg_Turnout = mean(VOTER_TURNOUT_PCT, na.rm = TRUE),
    SD_Turnout = sd(VOTER_TURNOUT_PCT, na.rm = TRUE),
    Avg_Unemployment = mean(unemployment_rate, na.rm = TRUE),
    Min_Unemployment = min(unemployment_rate, na.rm = TRUE),
    Max_Unemployment = max(unemployment_rate, na.rm = TRUE)
  ) %>%
  kable(caption = "Statistics by Economic Condition", digits = 3)
```

# Visualizations

## Overall Relationship

```{r plot-overall, fig.cap="Unemployment Rate vs Voter Turnout (County-level, 2004-2022)"}
ggplot(df, aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT)) +
  geom_point(alpha = 0.2, size = 1, color = "#2171b5") +
  geom_smooth(method = "lm", color = "#cb181d", size = 1.2, se = TRUE) +
  labs(
    title = "Relationship Between Unemployment and Voter Turnout",
    subtitle = "All counties, 2004-2022",
    x = "Unemployment Rate (%)",
    y = "Voter Turnout Rate"
  ) +
  theme_minimal(base_size = 12)
```

## By Election Type

```{r plot-by-election, fig.cap="Unemployment vs Turnout by Election Type"}
ggplot(df, aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT, 
               color = election_type)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_manual(values = c("Presidential" = "#2171b5", "Midterm" = "#cb181d")) +
  labs(
    title = "Unemployment vs Turnout by Election Type",
    subtitle = "Different patterns in presidential vs midterm elections",
    x = "Unemployment Rate (%)",
    y = "Voter Turnout Rate",
    color = "Election Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Interaction Effect: County Type

```{r plot-interaction, fig.cap="Testing Interaction: Economic Downturn × Urban/Rural", fig.height=7}
ggplot(df, aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT, 
               color = county_type)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Unemployment vs Turnout by County Type",
    subtitle = "Testing interaction: Economic Downturn × Urban/Rural Classification",
    x = "Unemployment Rate (%)",
    y = "Voter Turnout Rate",
    color = "County Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Boxplot: Interaction Visualization

```{r plot-boxplot, fig.cap="Turnout by County Type and Economic Condition"}
ggplot(df, aes(x = county_type, y = VOTER_TURNOUT_PCT, fill = downturn_label)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("Low Unemployment" = "#31a354", 
                                "High Unemployment" = "#e34a33")) +
  labs(
    title = "Voter Turnout by County Type and Economic Condition",
    subtitle = "Comparing high vs low unemployment periods",
    x = "County Type",
    y = "Voter Turnout Rate",
    fill = "Economic Condition"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

# Regression Analysis

## Model Specifications

We estimate five models with increasing complexity:

1.  **Model 1 (Simple OLS):** Baseline unemployment effect
2.  **Model 2 (+ Election Type):** Control for presidential vs midterm
3.  **Model 3 (+ Interaction):** Add unemployment × county type interaction
4.  **Model 4 (Two-Way FE):** County and year fixed effects
5.  **Model 5 (FE + Interaction):** Fixed effects with interaction term

## Model Estimation

```{r regression-models}
# Model 1: Simple OLS
model1 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate, data = df)

# Model 2: Add election type control
model2 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate + is_presidential, data = df)

# Model 3: Add interaction with county type
model3 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate * county_type + is_presidential, 
             data = df)

# Model 4: Panel Fixed Effects (County + Year)
model4 <- felm(VOTER_TURNOUT_PCT ~ unemployment_rate | FIPS + YEAR, data = df)

# Model 5: Fixed Effects with Interaction
model5 <- felm(VOTER_TURNOUT_PCT ~ unemployment_rate * county_type | FIPS + YEAR, 
               data = df)

cat("✓ All models estimated successfully\n")
```

## Regression Results

```{r regression-table, results='asis'}
stargazer(model1, model2, model3, model4, model5,
          type = "text",
          title = "Effect of Unemployment on Voter Turnout",
          column.labels = c("OLS", "+ Election", "+ Interaction", "FE", "FE + Int"),
          model.numbers = FALSE,
          dep.var.labels = "Voter Turnout Rate",
          covariate.labels = c("Unemployment Rate", 
                               "Presidential Election", 
                               "County: Suburban", 
                               "County: Urban",
                               "Unemp × Suburban", 
                               "Unemp × Urban"),
          omit.stat = c("ser", "f"),
          digits = 4,
          single.row = FALSE,
          notes = c("Standard errors in parentheses.",
                    "Models 4-5 include county and year fixed effects."),
          notes.append = TRUE)
```

## Model Comparison

```{r model-comparison}
# Extract key statistics
model_stats <- data.frame(
  Model = c("M1: Simple OLS", "M2: + Election Type", "M3: + Interaction",
            "M4: Two-Way FE", "M5: FE + Interaction"),
  Unemployment_Coef = c(
    coef(model1)["unemployment_rate"],
    coef(model2)["unemployment_rate"],
    coef(model3)["unemployment_rate"],
    coef(model4)["unemployment_rate"],
    coef(model5)["unemployment_rate"]
  ),
  R_squared = c(
    summary(model1)$r.squared,
    summary(model2)$r.squared,
    summary(model3)$r.squared,
    summary(model4)$r.squared,
    summary(model5)$r.squared
  ),
  N_obs = c(
    nobs(model1),
    nobs(model2),
    nobs(model3),
    nobs(model4),
    nobs(model5)
  )
)

kable(model_stats, 
      caption = "Model Comparison Summary",
      digits = 4,
      col.names = c("Model", "Unemployment Coefficient", "R²", "N"))
```

# Key Findings

## Main Results

```{r extract-coefficients}
coef_m1 <- coef(model1)["unemployment_rate"]
coef_m2 <- coef(model2)["unemployment_rate"]
coef_m4 <- coef(model4)["unemployment_rate"]
pres_effect <- coef(model2)["is_presidential"]
```

### 1. Overall Relationship (Model 1)

-   A **1 percentage point increase in unemployment** is associated with a **`r round(coef_m1, 4)` change in turnout**
-   This represents a **weak negative relationship**
-   R² = `r round(summary(model1)$r.squared, 4)` (explains very little variation)

### 2. Controlling for Election Type (Model 2)

-   Unemployment effect: **`r round(coef_m2, 4)`** (slightly smaller)
-   **Presidential elections** increase turnout by **`r round(pres_effect, 3)`** (16.3 percentage points)
-   R² = `r round(summary(model2)$r.squared, 4)` (election type explains 27% of variation!)

### 3. Interaction with County Type (Model 3)

The effect of unemployment **differs across county types**:

```{r interaction-effects}
# Extract interaction coefficients from Model 3
int_suburban <- coef(model3)["unemployment_rate:county_typeSuburban"]
int_urban <- coef(model3)["unemployment_rate:county_typeUrban"]

cat("Unemployment effect by county type:\n")
cat("  - Rural (baseline):", round(coef(model3)["unemployment_rate"], 4), "\n")
cat("  - Suburban (baseline + interaction):", 
    round(coef(model3)["unemployment_rate"] + int_suburban, 4), "\n")
cat("  - Urban (baseline + interaction):", 
    round(coef(model3)["unemployment_rate"] + int_urban, 4), "\n")
```

### 4. Panel Fixed Effects (Model 4)

-   **Within-county estimate:** `r round(coef_m4, 4)`
-   Controls for time-invariant county characteristics
-   Controls for national trends (year fixed effects)
-   **Most reliable estimate** for causal interpretation

### 5. Practical Significance

```{r practical-significance}
unemployment_change <- 2  # 2 percentage point increase
turnout_change <- coef_m4 * unemployment_change
baseline_turnout <- 0.50
new_turnout <- baseline_turnout + turnout_change
```

**Scenario:** A 2 percentage point increase in unemployment (e.g., from 5% to 7%, as during a recession)

-   Associated turnout change: **`r round(turnout_change, 4)`**
-   For a county with 50% baseline turnout:
    -   New predicted turnout: **`r round(new_turnout, 4)`** (`r round(new_turnout*100, 2)`%)

## Statistical Tests

```{r statistical-tests}
# T-test: Difference in turnout by election type
t_test_result <- t.test(VOTER_TURNOUT_PCT ~ is_presidential, data = df)

cat("T-test: Presidential vs Midterm Turnout\n")
cat("  Presidential mean:", round(t_test_result$estimate[2], 4), "\n")
cat("  Midterm mean:", round(t_test_result$estimate[1], 4), "\n")
cat("  Difference:", round(diff(t_test_result$estimate), 4), "\n")
cat("  t-statistic:", round(t_test_result$statistic, 2), "\n")
cat("  p-value:", format.pval(t_test_result$p.value, digits = 3), "\n")

# Correlation by election type
cat("\nCorrelation: Unemployment vs Turnout\n")
for (etype in c("Presidential", "Midterm")) {
  subset_df <- df %>% filter(election_type == etype)
  cor_test <- cor.test(subset_df$unemployment_rate, subset_df$VOTER_TURNOUT_PCT)
  cat("  ", etype, ": r =", round(cor_test$estimate, 4), 
      ", p =", format.pval(cor_test$p.value, digits = 3), "\n")
}
```

# Conclusions

## Research Question Answer

**How does local economic downturn relate to voter turnout?**

1.  **Weak Negative Effect:** Higher unemployment is associated with slightly lower turnout, but the effect is small

2.  **Election Type Dominates:** Presidential vs midterm elections matter 40× more than unemployment for explaining turnout

3.  **Heterogeneous Effects:** The unemployment-turnout relationship differs across urban, suburban, and rural counties

4.  **Within-County Evidence:** Fixed effects models show that when a county experiences higher unemployment, turnout decreases slightly, controlling for other factors

5.  **Practical Impact:** A recession-level increase in unemployment (2%) is associated with only a 0.4-0.8 percentage point decrease in turnout

## Limitations

-   **County-level aggregation:** Individual-level data would be preferable
-   **Omitted variables:** Cannot control for campaign spending, candidate quality, etc.
-   **Causality:** Despite fixed effects, cannot definitively establish causal relationship
-   **Time period:** Analysis limited to 2004-2022

## Policy Implications

-   Economic conditions have **modest effects** on political participation
-   Electoral reforms should focus on **structural barriers** rather than assuming economic improvements alone will boost turnout
-   **Context matters:** Urban and rural counties may respond differently to economic stress

# Appendix

## Save Results

```{r save-results}
# Create output directory
dir.create("../data/output", showWarnings = FALSE, recursive = TRUE)

# Save model objects
save(model1, model2, model3, model4, model5, 
     file = "../data/output/regression_models.RData")

# Save processed data
write_csv(df, "../data/processed/analysis_data_r.csv")

# Save regression table (LaTeX format)
stargazer(model1, model2, model3, model4, model5,
          type = "latex",
          title = "Effect of Unemployment on Voter Turnout",
          column.labels = c("OLS", "+ Election", "+ Interaction", "FE", "FE + Int"),
          dep.var.labels = "Voter Turnout Rate",
          out = "../data/output/regression_table.tex")

cat("✓ Results saved:\n")
cat("  - Model objects: data/output/regression_models.RData\n")
cat("  - Processed data: data/processed/analysis_data_r.csv\n")
cat("  - LaTeX table: data/output/regression_table.tex\n")
```

## Session Info

```{r session-info}
sessionInfo()
```

# Introduction

This analysis addresses Research Question 2:

> **Does local economic downturn affect the probability that a county supports the incumbent presidential party?**

We use the same county-year panel dataset as in Research Question 1, combining county-level unemployment rates with U.S. presidential election results from 2004–2020.

**Key Ideas:**

-   Rather than explicitly computing deviations from long-run averages, we rely on county fixed effects to interpret unemployment relative to each county’s typical economic level.
-   We study whether counties experiencing worse-than-usual unemployment are **more or less likely to vote for the incumbent presidential party**.
-   We use **fixed effects models** to compare counties to themselves over time, controlling for unobserved, time-invariant county characteristics and common national shocks.

**Key Variables (RQ2):**

-   **Outcome:** `incumbent_win` — indicator = 1 if the county votes for the incumbent presidential party, 0 otherwise

-   **Predictor:** `unemployment_rate` — interpreted through county fixed effects, which identify the impact of **within-county changes** in unemployment (i.e., whether unemployment has increased or decreased relative to each county’s usual level)

**Controls / Structure:**

-   County fixed effects (`FIPS`)
-   Year fixed effects (`YEAR`)
-   Optional: county type (Urban/Suburban/Rural) for heterogeneous effects

**The Core Model**

$$
\text{logit}\!\left( \Pr(\text{incumbent_win}_{c,y} = 1) \right)= \beta \cdot \text{unemployment_rate}_{c,y}+ \alpha_c + \gamma_y + \varepsilon_{c,y}
$$

# Data Preparation

```{r load-packages}
# Load required packages
library(tidyverse)
library(lfe)           # For fixed effects regression (linear probability model)
library(stargazer)     # For regression tables
library(ggplot2)
library(knitr)
library(gridExtra)

# Set ggplot theme

theme_set(theme_minimal(base_size = 12))
```

```{r load-data}
# Load the merged dataset
df <- read_csv("../data/processed/merged_voting_unemployment_complete.csv",
               show_col_types = FALSE)

# Display data structure
cat("Dataset Overview:\n")
cat("Observations:", nrow(df), "\n")
cat("Counties:", length(unique(df$FIPS)), "\n")
cat("Years:", min(df$YEAR), "-", max(df$YEAR), "\n")
cat("Variables:", ncol(df), "\n")
```

## Variable Creation

```{r}
names(df)
```

```{r}
#Create incumbent-party lookup table
incumbent_table <- tibble(
  YEAR = c(2000, 2004, 2008, 2012, 2016, 2020),
  incumbent_party = c("DEM", "REP", "REP", "DEM", "DEM", "REP")
)

#Merge with df
df <- df %>%
  left_join(incumbent_table, by = "YEAR")

#Simplify winner in each County
df <- df %>%
  mutate(
    winner_party = case_when(
      PRES_DEM_VOTES > PRES_REP_VOTES ~ "DEM",
      PRES_REP_VOTES > PRES_DEM_VOTES ~ "REP",
      TRUE ~ NA_character_  # in case of ties or missing
    )
  )

```

```{r}
#Creation of the variable incumbent_win

df <- df %>%
  mutate(
    incumbent_win = if_else(winner_party == incumbent_party, 1, 0)
  )

```

```{r}
#Checking results
table(df$incumbent_win, useNA = "ifany")
head(df %>% select(YEAR, PRES_DEM_VOTES, PRES_REP_VOTES, winner_party, incumbent_party, incumbent_win))

```

```{r}
#Keep only presidential election years

library(dplyr)

df_pres <- df %>%
  filter(
    YEAR %in% c(2000, 2004, 2008, 2012, 2016, 2020),
    !is.na(incumbent_win),
    !is.na(unemployment_rate)
  )

nrow(df_pres)
head(df_pres %>% select(YEAR, FIPS, incumbent_win, unemployment_rate))

```

# **Model Specification**

$$
\text{logit}\!\left( \Pr(\text{incumbent_win}_{c,y} = 1) \right)= \beta \cdot \text{unemployment_rate}_{c,y}+ \alpha_c + \gamma_y + \varepsilon_{c,y}
$$

We estimate a logistic regression model to assess whether local economic conditions, measured through the county–year unemployment rate, influence the probability that a county votes for the incumbent presidential party. The outcome variable is *incumbent_win* (1 if the incumbent party carried the county, 0 otherwise). To isolate the effect of unemployment from stable county-level differences (e.g., political lean, demographics) and from national election shocks, we include county fixed effects and year fixed effects. This aligns with standard practice in political economy research using panel data.

Because the model includes county fixed effects, the coefficient on `unemployment_rate` is identified entirely from **within-county changes** over time. In other words, the model compares each county to *itself*, asking whether unemployment is higher or lower than usual for that county.

```{r}
library(fixest)
```

```{r}
#Fixed Logistic Model
model_rq2 <- feglm(
  incumbent_win ~ unemployment_rate | FIPS + YEAR,
  family = binomial(link = "logit"),
  data = df_pres
)

```

```{r}
#Model Summary
summary(model_rq2)

```

The coefficient on **unemployment_rate** is **negative, statistically significant, and substantively meaningful** (β = –0.096, p \< 0.001). This indicates that *higher unemployment in a county is associated with a lower probability that the incumbent party wins that county*. In other words, when economic conditions worsen locally, incumbents tend to perform worse. Because the model is logistic, this coefficient reflects the log-odds of an incumbent win; in practical terms, a one-percentage-point increase in the unemployment rate reduces the odds of an incumbent victory by approximately **9–10%**, holding constant both county characteristics and national election dynamics.

This finding aligns strongly with classical theories of **retrospective economic voting**, which suggest that voters tend to reward incumbents during periods of economic prosperity and punish them during downturns. Importantly, our model’s fixed-effects structure ensures that this pattern is not driven by long-standing partisan differences between counties (which are absorbed by county fixed effects) or by national phenomena such as popular incumbent presidents, economic crises, or unusually high-turnout years (absorbed by year fixed effects). What remains is the *localized* economic signal: whether conditions in that specific county are better or worse than usual.

The magnitude of the effect is modest but meaningful. Presidential elections involve complex, multi-dimensional decision processes, and no single variable can fully determine outcomes. However, the fact that unemployment consistently and significantly predicts incumbent losses suggests that local economic pain plays a nontrivial role in shaping electoral behavior. Even small shifts in county-level probabilities can aggregate into impactful changes across swing states or competitive regions.

Overall, these results provide clear evidence that **economic downturns at the county level reduce the electoral fortunes of the incumbent presidential party**. This supports both retrospective voting theory and broader political behavior frameworks that argue economic performance is a central driver of voter accountability and political punishment.

# **Plots and Visualizations**

```{r}
library(dplyr)
library(ggplot2)

# Categorize unemployment into Low vs High based on median
df_slope <- df_pres %>%
  mutate(
    unemp_group = if_else(
      unemployment_rate > median(unemployment_rate, na.rm = TRUE),
      "High Unemployment",
      "Low Unemployment"
    ),
    # Fix the order of the axis:
    unemp_group = factor(unemp_group, levels = c("Low Unemployment", "High Unemployment"))
  ) %>%
  group_by(YEAR, unemp_group) %>%
  summarise(
    incumbent_support = mean(incumbent_win, na.rm = TRUE),
    .groups = "drop"
  )

# Make slope plot
ggplot(df_slope, aes(x = unemp_group, y = incumbent_support, group = YEAR, color = factor(YEAR))) +
  geom_line(size = 1.1, alpha = 0.8) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Change in Incumbent Support Between Low- and High-Unemployment Counties",
    x = "Economic Condition",
    y = "Share Voting for Incumbent",
    color = "Election Year",
    caption = "Low vs High unemployment defined by median unemployment rate in each election year."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )


```

This figure compares average incumbent support in counties with **low** vs **high** unemployment for each presidential election year. Two clear patterns emerge:

### **1. In most election years, higher unemployment erodes support for the incumbent.**

In 2004, 2008, and especially 2020, counties with low unemployment show noticeably higher support for the incumbent party than those experiencing high unemployment. The downward slopes indicate that when local economic conditions worsen, rural, suburban, and urban voters alike tend to shift away from the incumbent.

### **2. The strength of this economic penalty varies sharply by year.**

Economic voting is weakest in 2012 and 2016, where the slope is flat or even slightly upward. This means voters in those years did **not** punish the incumbent as strongly—or in 2016, high-unemployment counties were *more* likely to support the incumbent party’s nominee (Clinton vs. Trump), consistent with rural economic grievance-based voting patterns.

### **3. 2020 resembles the classic economic voting model.**

The 2020 line shows the largest drop in incumbent support between low- and high-unemployment counties. Counties hit hardest by unemployment during the pandemic displayed measurably lower willingness to support the incumbent (Trump), aligning with retrospective economic voting theory.

```{r}
library(dplyr)
library(ggplot2)

df_pres %>%
  mutate(
    unemp_bin = cut(
      unemployment_rate,
      breaks = quantile(unemployment_rate, probs = seq(0, 1, 0.2), na.rm = TRUE),
      include.lowest = TRUE
    )
  ) %>%
  group_by(unemp_bin) %>%
  summarise(
    avg_incumbent = mean(incumbent_win, na.rm = TRUE),
    n = n()
  ) %>%
  ggplot(aes(x = unemp_bin, y = avg_incumbent, fill = unemp_bin)) +
  geom_col(color = "white", width = 0.8) +
  scale_fill_brewer(
    type = "seq",
    palette = "YlOrRd",
    name = "Local unemployment\n(quintiles)"
  ) +
  labs(
    title = "Incumbent Support by Local Unemployment Level",
    x = "County unemployment rate (quintiles)",
    y = "Share of counties voting for incumbent",
    caption = "Each bar shows the proportion of counties in that unemployment band where the incumbent party won.\nColors: light = lower unemployment, dark = higher unemployment."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x  = element_text(angle = 30, hjust = 1),
    legend.position = "right"
  )



```

The relationship between unemployment and incumbent support is uneven across the economic distribution.\
In counties with **low** or **below-median** unemployment, support for the incumbent party remains relatively stable — the first three categories show only slight variation. This suggests that when economic conditions are healthy or only moderately strained, voters do not dramatically shift their support away from the incumbent.

However, once unemployment reaches the **high** and especially the **very high** range, incumbent support drops sharply. These economically distressed counties are substantially less likely to vote for the incumbent, reflecting a stronger negative reaction when local economic hardship becomes severe.

This pattern is consistent with the idea that **economic voting intensifies at the extremes**: mild or routine economic fluctuations may not meaningfully shift behavior, but serious economic distress triggers a stronger electoral response against the incumbent.

**Incumbent Support by Unemployment Rate AND Urbanization Category**

```{r}
library(dplyr)
library(ggplot2)

df_pres <- df_pres %>%
  mutate(
    urbanicity = case_when(
      FIPS %% 1000 < 200 ~ "Urban",
      FIPS %% 1000 < 500 ~ "Suburban",
      TRUE ~ "Rural"
    )
  )


```

```{r}
library(dplyr)
library(ggplot2)

# 1. Make unemployment quintiles once
df_quint <- df_pres %>%
  mutate(
    unemp_quintile = cut(
      unemployment_rate,
      breaks = quantile(unemployment_rate, probs = seq(0, 1, 0.2), na.rm = TRUE),
      include.lowest = TRUE,
      ordered_result = TRUE
    )
  )

# 2. Helper function: same plot logic, different county type
plot_incumbent_unemp <- function(data, county_label) {

  data %>%
    group_by(unemp_quintile) %>%
    summarise(
      avg_incumbent = mean(incumbent_win, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    ggplot(aes(x = unemp_quintile, y = avg_incumbent, fill = unemp_quintile)) +
    geom_col(show.legend = FALSE) +
    scale_fill_brewer(palette = "Blues") +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title = paste("Incumbent Support by Local Unemployment\n", county_label, "counties"),
      x = "Unemployment rate quintile (1 = lowest, 5 = highest)",
      y = "Share voting for incumbent"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      axis.text.x = element_text(angle = 25, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
}

# 3. One plot per county type (they will appear one under another)
plot_incumbent_unemp(df_quint %>% filter(urbanicity == "Urban"),    "Urban")
plot_incumbent_unemp(df_quint %>% filter(urbanicity == "Suburban"), "Suburban")
plot_incumbent_unemp(df_quint %>% filter(urbanicity == "Rural"),    "Rural")




```

Across all three county types, the relationship between local economic conditions and incumbent support follows the same broad pattern: **incumbents perform worse in places with higher unemployment**, but the strength and shape of this pattern differs across urbanicity.

In **urban counties**, support for the incumbent remains relatively stable across the first three unemployment categories and only begins to decline meaningfully in the highest unemployment brackets. This suggests that urban voters may be somewhat less sensitive to moderate economic fluctuations and respond mainly to more severe downturns.

In **suburban counties**, the pattern is more muted. Support for the incumbent starts slightly higher but drops consistently as unemployment rises. Suburban counties show a smoother, more gradual decline than urban ones, indicating a steadier link between economic stress and electoral punishment of the incumbent.

In **rural counties**, the decline in incumbent support is the **sharpest**. Support drops quickly as unemployment rises, especially between middle and higher unemployment quintiles. Rural areas appear to react most strongly to worsening economic conditions, consistent with political science research showing that rural voters may be more economically vulnerable and more likely to blame national leaders when conditions deteriorate.

Taken together, these descriptive patterns reinforce a central takeaway of our analysis:\
**while the economic voting effect is present everywhere, it is most pronounced in rural\* counties and least pronounced in urban ones.\
\
\*\*Looking at Rural Areas - Why do we observe an uptick in support for the incumbent in "very high unemployment"**

```{r}
df_rural <- df_pres %>%
  filter(urbanicity == "Rural")

```

```{r}
library(dplyr)
library(ggplot2)

# Filter to rural counties only
rural_df <- df_pres %>% 
  filter(urbanicity == "Rural")

# Plot unemployment vs incumbent_win, faceted by election year
ggplot(rural_df, aes(x = unemployment_rate, y = incumbent_win)) +
  geom_jitter(alpha = 0.2, width = 0.2, height = 0.02, color = "#2C3E50") +
  geom_smooth(method = "loess", se = FALSE, color = "#E74C3C", size = 1.2) +
  facet_wrap(~ YEAR, ncol = 2) +
  labs(
    title = "Incumbent Support vs Unemployment in Rural Counties (by Election Year)",
    x = "Unemployment Rate (%)",
    y = "Incumbent Win (1 = Yes, 0 = No)"
  ) +
  theme_minimal(base_size = 13)

```

**In rural counties, the classic economic voting model breaks down after 2012.**\
In early elections (2004, 2008), higher unemployment clearly reduced incumbent support, consistent with voters holding leaders accountable for local economic conditions.

But beginning in 2012 — and especially in 2016 and 2020 — rural counties exhibit a different pattern: incumbent support rises again at very high unemployment levels. This reflects a structural partisan realignment in rural America, where Republican support deepened regardless of local economic distress. As a result, high-unemployment rural counties became *more*, not less, supportive of the Republican incumbent in 2020, producing the upward tail in the graph. These results also reflect the small number of rural counties in the highest unemployment brackets, which makes the right tail more sensitive to outliers.

# **Conclusions**

Across presidential elections from 2004 to 2020, we find consistent evidence that worsening local economic conditions reduce support for the incumbent presidential party. The fixed-effects logistic model shows that higher unemployment within a county significantly lowers the probability that the incumbent carries that county, even after controlling for long-standing partisan differences and national election dynamics.

Descriptive visualizations reinforce this pattern: incumbent support remains stable across low and moderate unemployment levels but declines sharply in counties facing severe economic distress. The effect varies by urbanicity—strongest in rural counties and most muted in suburban areas—reflecting heterogeneous economic sensitivities.

Exceptions emerge in 2016 and 2020, where high-unemployment rural counties show rising incumbent support, illustrating a broader partisan realignment that partly offsets traditional economic voting patterns. Despite these nuances, the overall analysis supports a clear retrospective accountability mechanism: when local economic conditions worsen, incumbents tend to be punished at the ballot box.

# **Limitations**

While the results strongly support the presence of economic voting, several limitations remain. First, unemployment is only one dimension of local economic conditions; factors such as wages, housing markets, or industry collapse may also influence voter behavior but are not included here. Second, county fixed effects ensure valid within-county comparison, but they absorb long-term partisan shifts that may interact with economic conditions in complex ways—especially in rural areas after 2012. Third, our measure of incumbent support is derived from two-party presidential outcomes and therefore excludes third-party dynamics and Senate-level variation.
