---
title: "Economic Downturn and Incumbent Support"
subtitle: "Research Question 2: Does local economic downturn affect support for the incumbent presidential party?"
author: "Statistical Analysis - STA 702"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    self-contained: true
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# Introduction

This analysis addresses Research Question 2:

> **Does local economic downturn affect the probability that a county supports the incumbent presidential party?**

We use the same county-year panel dataset as in Research Question 1, combining county-level unemployment rates with U.S. presidential election results from 2004–2020.

**Key Ideas:**

-   Rather than using unemployment *levels*, we focus on **unemployment relative to a county’s own long-run average** (local economic stress).
-   We study whether counties experiencing worse-than-usual unemployment are **more or less likely to vote for the incumbent presidential party**.
-   We use **fixed effects models** to compare counties to themselves over time, controlling for unobserved, time-invariant county characteristics and common national shocks.

**Key Variables (RQ2):**

-   **Outcome:** `incumbent_win` — indicator = 1 if the county votes for the incumbent presidential party, 0 otherwise\

-   **Predictor:** `unemployment_rate` — interpreted through county fixed effects, which identify the impact of **within-county changes** in unemployment (i.e., whether unemployment has increased or decreased relative to each county’s usual level)

**Controls / Structure:**

-   County fixed effects (`FIPS`)\
-   Year fixed effects (`YEAR`)\
-   Optional: county type (Urban/Suburban/Rural) for heterogeneous effects

**The Core Model**

$$
\text{logit}\!\left( \Pr(\text{incumbent_win}_{c,y} = 1) \right)= \beta \cdot \text{unemployment_rate}_{c,y}+ \alpha_c + \gamma_y + \varepsilon_{c,y}
$$

# Data Preparation

```{r load-packages}
# Load required packages
library(tidyverse)
library(lfe)           # For fixed effects regression (linear probability model)
library(stargazer)     # For regression tables
library(ggplot2)
library(knitr)
library(gridExtra)

# Set ggplot theme

theme_set(theme_minimal(base_size = 12))
```

```{r load-data}
# Load the merged dataset
df <- read_csv("../data/processed/merged_voting_unemployment_complete.csv",
               show_col_types = FALSE)

# Display data structure
cat("Dataset Overview:\n")
cat("Observations:", nrow(df), "\n")
cat("Counties:", length(unique(df$FIPS)), "\n")
cat("Years:", min(df$YEAR), "-", max(df$YEAR), "\n")
cat("Variables:", ncol(df), "\n")
```

## Variable Creation

```{r}
names(df)
```

```{r}
#Create incumbent-party lookup table
incumbent_table <- tibble(
  YEAR = c(2000, 2004, 2008, 2012, 2016, 2020),
  incumbent_party = c("DEM", "REP", "REP", "DEM", "DEM", "REP")
)

#Merge with df
df <- df %>%
  left_join(incumbent_table, by = "YEAR")

#Simplify winner in each County
df <- df %>%
  mutate(
    winner_party = case_when(
      PRES_DEM_VOTES > PRES_REP_VOTES ~ "DEM",
      PRES_REP_VOTES > PRES_DEM_VOTES ~ "REP",
      TRUE ~ NA_character_  # in case of ties or missing
    )
  )

```

```{r}
#Creation of the variable incumbent_win

df <- df %>%
  mutate(
    incumbent_win = if_else(winner_party == incumbent_party, 1, 0)
  )

```

```{r}
#Checking results
table(df$incumbent_win, useNA = "ifany")
head(df %>% select(YEAR, PRES_DEM_VOTES, PRES_REP_VOTES, winner_party, incumbent_party, incumbent_win))

```

```{r}
#Keep only presidential election years

library(dplyr)

df_pres <- df %>%
  filter(
    YEAR %in% c(2000, 2004, 2008, 2012, 2016, 2020),
    !is.na(incumbent_win),
    !is.na(unemployment_rate)
  )

nrow(df_pres)
head(df_pres %>% select(YEAR, FIPS, incumbent_win, unemployment_rate))

```

# **Model Specification**

$$
\text{logit}\!\left( \Pr(\text{incumbent_win}_{c,y} = 1) \right)= \beta \cdot \text{unemployment_rate}_{c,y}+ \alpha_c + \gamma_y + \varepsilon_{c,y}
$$

We estimate a logistic regression model to assess whether local economic conditions, measured through the county–year unemployment rate, influence the probability that a county votes for the incumbent presidential party. The outcome variable is *incumbent_win* (1 if the incumbent party carried the county, 0 otherwise). To isolate the effect of unemployment from stable county-level differences (e.g., political lean, demographics) and from national election shocks, we include county fixed effects and year fixed effects. This aligns with standard practice in political economy research using panel data.

Because the model includes county fixed effects, the coefficient on `unemployment_rate` is identified entirely from **within-county changes** over time. In other words, the model compares each county to *itself*, asking whether unemployment is higher or lower than usual for that county. This aligns directly with Tea’s requirement to evaluate whether local unemployment has “increased or decreased” relative to each county’s baseline economic conditions.

```{r}
library(fixest)
```

```{r}
#Fixed Logistic Model
model_rq2 <- feglm(
  incumbent_win ~ unemployment_rate | FIPS + YEAR,
  family = binomial(link = "logit"),
  data = df_pres
)

```

```{r}
#Model Summary
summary(model_rq2)

```

The coefficient on **unemployment_rate** is **negative, statistically significant, and substantively meaningful** (β = –0.096, p \< 0.001). This indicates that *higher unemployment in a county is associated with a lower probability that the incumbent party wins that county*. In other words, when economic conditions worsen locally, incumbents tend to perform worse. Because the model is logistic, this coefficient reflects the log-odds of an incumbent win; in practical terms, a one-percentage-point increase in the unemployment rate reduces the odds of an incumbent victory by approximately **9–10%**, holding constant both county characteristics and national election dynamics.

This finding aligns strongly with classical theories of **retrospective economic voting**, which suggest that voters tend to reward incumbents during periods of economic prosperity and punish them during downturns. Importantly, our model’s fixed-effects structure ensures that this pattern is not driven by long-standing partisan differences between counties (which are absorbed by county fixed effects) or by national phenomena such as popular incumbent presidents, economic crises, or unusually high-turnout years (absorbed by year fixed effects). What remains is the *localized* economic signal: whether conditions in that specific county are better or worse than usual.

The magnitude of the effect is modest but meaningful. Presidential elections involve complex, multi-dimensional decision processes, and no single variable can fully determine outcomes. However, the fact that unemployment consistently and significantly predicts incumbent losses suggests that local economic pain plays a nontrivial role in shaping electoral behavior. Even small shifts in county-level probabilities can aggregate into impactful changes across swing states or competitive regions.

Overall, these results provide clear evidence that **economic downturns at the county level reduce the electoral fortunes of the incumbent presidential party**. This supports both retrospective voting theory and broader political behavior frameworks that argue economic performance is a central driver of voter accountability and political punishment.

# **Plots and Visualizations**

**Predicted Probability of Incumbent Win by Unemployment Rate**

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

base_row <- df_pres %>%
  filter(!is.na(FIPS), !is.na(YEAR)) %>%
  slice(1) %>%                  
  select(FIPS, YEAR)          

df_pred <- tibble(
  unemployment_rate = seq(
    min(df_pres$unemployment_rate, na.rm = TRUE),
    max(df_pres$unemployment_rate, na.rm = TRUE),
    length.out = 200
  )
) %>%
  crossing(base_row)

df_pred$pred <- predict(
  model_rq2,
  newdata = df_pred,
  type = "response"
)

ggplot(df_pred, aes(x = unemployment_rate, y = pred)) +
  geom_line(size = 1.2) +
  labs(
    title = "Predicted Probability of Incumbent Win by Unemployment Rate",
    x = "Unemployment Rate (%)",
    y = "Predicted Probability")
```

This figure shows the model-based relationship between county unemployment rates and the probability of an incumbent presidential victory. Consistent with the regression results, the predicted probability declines as unemployment increases. Counties with relatively low unemployment (around 3–5%) are estimated to have a probability of incumbent support above 0.85, while counties facing unemployment levels above 15–20% show substantially lower predicted support.

The smooth downward curve reflects the logistic functional form and the negative coefficient on unemployment in the fixed-effects model. Overall, this visualization provides clear evidence that higher unemployment is associated with reduced electoral support for the incumbent presidential party.

**Incumbent Support by Unemployment Rate Quintiles**

```{r}
df_pres %>%
  mutate(
    unemp_bin = cut(
      unemployment_rate,
      breaks = quantile(unemployment_rate, probs = seq(0, 1, 0.2), na.rm = TRUE),
      include.lowest = TRUE
    )
  ) %>%
  group_by(unemp_bin) %>%
  summarise(
    avg_incumbent = mean(incumbent_win, na.rm = TRUE),
    n = n()
  ) %>%
  ggplot(aes(x = unemp_bin, y = avg_incumbent)) +
  geom_col(fill = "#2C3E50", alpha = 0.8) +
  labs(
    title = "Incumbent Support by Unemployment Rate Quintiles",
    x = "Unemployment Rate (Quintiles)",
    y = "Share Voting for Incumbent"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```

This figure summarizes the average share of votes for the incumbent party across unemployment-rate quintiles. Counties in the lowest unemployment quintiles exhibit higher incumbent support, typically around 0.55–0.60. In contrast, counties in the highest unemployment quintile show markedly lower incumbent support, declining to roughly 0.47.

The pattern is consistent with the regression results: incumbent support falls most noticeably in counties experiencing the most severe economic conditions. This descriptive comparison reinforces the main finding that local economic stress corresponds to weaker support for the incumbent.

**Incumbent Support in High vs Low Unemployment Counties**

```{r}
df_pres %>%
  mutate(
    high_unemp = if_else(
      unemployment_rate > median(unemployment_rate, na.rm = TRUE),
      "High Unemployment",
      "Low Unemployment"
    )
  ) %>%
  group_by(high_unemp) %>%
  summarise(
    avg_incumbent = mean(incumbent_win, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = high_unemp, y = avg_incumbent, fill = high_unemp)) +
  geom_col(alpha = 0.85) +
  guides(fill = "none") +
  labs(
    title = "Incumbent Support in High vs Low Unemployment Counties",
    x = "Economic Condition",
    y = "Share Voting for Incumbent"
  ) +
  theme_minimal(base_size = 13)

```

This plot compares average incumbent support between counties above and below the median unemployment rate. Counties with lower unemployment have a higher probability of supporting the incumbent, while those experiencing above-median unemployment show reduced incumbent support.

Although this visualization is descriptive, it aligns closely with the fixed-effects estimates, indicating that incumbent support tends to decline when local economic conditions deteriorate. The difference between the two groups provides an intuitive summary of the economic voting pattern found in the analysis.
