---
title: "Unemployment and Voter Turnout Analysis"
subtitle: "Research Question 1: How does local economic downturn relate to voter turnout?"
author: "Statistical Analysis - STA 702"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    self-contained: true
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# Introduction

This analysis investigates the relationship between local economic conditions (measured by unemployment rate) and voter turnout in U.S. presidential and midterm elections from 2004-2022.

**Research Question:** How does local economic downturn relate to voter turnout in U.S. elections at the county level?

**Key Variables:**

-   **Outcome:** `VOTER_TURNOUT_PCT` = Voter turnout rate (continuous)
-   **Predictor:** `unemployment_rate` = County unemployment rate (%)
-   **Interaction:** `unemployment_rate × county_type` (Urban/Rural classification)

# Data Preparation

```{r load-packages}
# Load required packages
library(tidyverse)
library(lfe)           # For fixed effects regression
library(stargazer)     # For regression tables
library(ggplot2)
library(knitr)
library(gridExtra)

# Set ggplot theme
theme_set(theme_minimal(base_size = 12))
```

```{r load-data}
# Load the merged dataset
df <- read_csv("../data/processed/merged_voting_unemployment_complete.csv",
               show_col_types = FALSE)

# Display data structure
cat("Dataset Overview:\n")
cat("Observations:", nrow(df), "\n")
cat("Counties:", length(unique(df$FIPS)), "\n")
cat("Years:", min(df$YEAR), "-", max(df$YEAR), "\n")
cat("Variables:", ncol(df), "\n")
```

## Variable Creation

```{r create-variables}
df <- df %>%
  mutate(
    # Election type indicator
    is_presidential = ifelse(YEAR %% 4 == 0, 1, 0),
    election_type = ifelse(is_presidential == 1, "Presidential", "Midterm"),
    
    # Economic downturn indicator (unemployment > median)
    median_unemp = median(unemployment_rate, na.rm = TRUE),
    economic_downturn = ifelse(unemployment_rate > median_unemp, 1, 0),
    downturn_label = ifelse(economic_downturn == 1, 
                           "High Unemployment", "Low Unemployment"),
    
    # Standardized variables
    unemployment_std = (unemployment_rate - mean(unemployment_rate)) / 
                       sd(unemployment_rate),
    
    # Urban/rural classification based on CVAP
    # Large counties (>50k CVAP) = Urban
    # Medium counties (10k-50k) = Suburban  
    # Small counties (<10k) = Rural
    county_type = case_when(
      CVAP > 50000 ~ "Urban",
      CVAP > 10000 ~ "Suburban",
      TRUE ~ "Rural"
    ),
    
    # Factor variables for regression
    FIPS_factor = as.factor(FIPS),
    YEAR_factor = as.factor(YEAR),
    STATE_factor = as.factor(STATE_FIPS)
  )

cat("✓ Created classification variables\n")
cat("  - Median unemployment:", round(df$median_unemp[1], 2), "%\n")
```

# Descriptive Statistics

## Summary Statistics

```{r summary-stats}
# Overall summary
df %>%
  select(VOTER_TURNOUT_PCT, unemployment_rate, CVAP) %>%
  summary() %>%
  kable(caption = "Summary Statistics: Key Variables", digits = 3)
```

## Distribution by Groups

```{r stats-by-group}
# By election type
df %>%
  group_by(election_type) %>%
  summarise(
    N = n(),
    Avg_Turnout = mean(VOTER_TURNOUT_PCT, na.rm = TRUE),
    SD_Turnout = sd(VOTER_TURNOUT_PCT, na.rm = TRUE),
    Avg_Unemployment = mean(unemployment_rate, na.rm = TRUE),
    SD_Unemployment = sd(unemployment_rate, na.rm = TRUE)
  ) %>%
  kable(caption = "Statistics by Election Type", digits = 3)

# By county type
df %>%
  group_by(county_type) %>%
  summarise(
    N = n(),
    Counties = n_distinct(FIPS),
    Avg_Turnout = mean(VOTER_TURNOUT_PCT, na.rm = TRUE),
    SD_Turnout = sd(VOTER_TURNOUT_PCT, na.rm = TRUE),
    Avg_Unemployment = mean(unemployment_rate, na.rm = TRUE),
    SD_Unemployment = sd(unemployment_rate, na.rm = TRUE)
  ) %>%
  kable(caption = "Statistics by County Type", digits = 3)

# By economic downturn status
df %>%
  group_by(downturn_label) %>%
  summarise(
    N = n(),
    Avg_Turnout = mean(VOTER_TURNOUT_PCT, na.rm = TRUE),
    SD_Turnout = sd(VOTER_TURNOUT_PCT, na.rm = TRUE),
    Avg_Unemployment = mean(unemployment_rate, na.rm = TRUE),
    Min_Unemployment = min(unemployment_rate, na.rm = TRUE),
    Max_Unemployment = max(unemployment_rate, na.rm = TRUE)
  ) %>%
  kable(caption = "Statistics by Economic Condition", digits = 3)
```

# Visualizations

## Overall Relationship

```{r plot-overall, fig.cap="Unemployment Rate vs Voter Turnout (County-level, 2004-2022)"}
ggplot(df, aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT)) +
  geom_point(alpha = 0.2, size = 1, color = "#2171b5") +
  geom_smooth(method = "lm", color = "#cb181d", size = 1.2, se = TRUE) +
  labs(
    title = "Relationship Between Unemployment and Voter Turnout",
    subtitle = "All counties, 2004-2022",
    x = "Unemployment Rate (%)",
    y = "Voter Turnout Rate"
  ) +
  theme_minimal(base_size = 12)
```

## By Election Type

```{r plot-by-election, fig.cap="Unemployment vs Turnout by Election Type"}
ggplot(df, aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT, 
               color = election_type)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_manual(values = c("Presidential" = "#2171b5", "Midterm" = "#cb181d")) +
  labs(
    title = "Unemployment vs Turnout by Election Type",
    subtitle = "Different patterns in presidential vs midterm elections",
    x = "Unemployment Rate (%)",
    y = "Voter Turnout Rate",
    color = "Election Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Interaction Effect: County Type

```{r plot-interaction, fig.cap="Testing Interaction: Economic Downturn × Urban/Rural", fig.height=7}
ggplot(df, aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT, 
               color = county_type)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Unemployment vs Turnout by County Type",
    subtitle = "Testing interaction: Economic Downturn × Urban/Rural Classification",
    x = "Unemployment Rate (%)",
    y = "Voter Turnout Rate",
    color = "County Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Boxplot: Interaction Visualization

```{r plot-boxplot, fig.cap="Turnout by County Type and Economic Condition"}
ggplot(df, aes(x = county_type, y = VOTER_TURNOUT_PCT, fill = downturn_label)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("Low Unemployment" = "#31a354", 
                                "High Unemployment" = "#e34a33")) +
  labs(
    title = "Voter Turnout by County Type and Economic Condition",
    subtitle = "Comparing high vs low unemployment periods",
    x = "County Type",
    y = "Voter Turnout Rate",
    fill = "Economic Condition"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

# Regression Analysis

## Model Specifications

We estimate five models with increasing complexity:

1.  **Model 1 (Simple OLS):** Baseline unemployment effect
2.  **Model 2 (+ Election Type):** Control for presidential vs midterm
3.  **Model 3 (+ Interaction):** Add unemployment × county type interaction
4.  **Model 4 (Two-Way FE):** County and year fixed effects
5.  **Model 5 (FE + Interaction):** Fixed effects with interaction term

## Model Estimation

```{r regression-models}
# Model 1: Simple OLS
model1 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate, data = df)

# Model 2: Add election type control
model2 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate + is_presidential, data = df)

# Model 3: Add interaction with county type
model3 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate * county_type + is_presidential, 
             data = df)

# Model 4: Panel Fixed Effects (County + Year)
model4 <- felm(VOTER_TURNOUT_PCT ~ unemployment_rate | FIPS + YEAR, data = df)

# Model 5: Fixed Effects with Interaction
model5 <- felm(VOTER_TURNOUT_PCT ~ unemployment_rate * county_type | FIPS + YEAR, 
               data = df)

cat("✓ All models estimated successfully\n")
```

## Regression Results

```{r regression-table, results='asis'}
stargazer(model1, model2, model3, model4, model5,
          type = "text",
          title = "Effect of Unemployment on Voter Turnout",
          column.labels = c("OLS", "+ Election", "+ Interaction", "FE", "FE + Int"),
          model.numbers = FALSE,
          dep.var.labels = "Voter Turnout Rate",
          covariate.labels = c("Unemployment Rate", 
                               "Presidential Election", 
                               "County: Suburban", 
                               "County: Urban",
                               "Unemp × Suburban", 
                               "Unemp × Urban"),
          omit.stat = c("ser", "f"),
          digits = 4,
          single.row = FALSE,
          notes = c("Standard errors in parentheses.",
                    "Models 4-5 include county and year fixed effects."),
          notes.append = TRUE)
```

## Model Comparison

```{r model-comparison}
# Extract key statistics
model_stats <- data.frame(
  Model = c("M1: Simple OLS", "M2: + Election Type", "M3: + Interaction",
            "M4: Two-Way FE", "M5: FE + Interaction"),
  Unemployment_Coef = c(
    coef(model1)["unemployment_rate"],
    coef(model2)["unemployment_rate"],
    coef(model3)["unemployment_rate"],
    coef(model4)["unemployment_rate"],
    coef(model5)["unemployment_rate"]
  ),
  R_squared = c(
    summary(model1)$r.squared,
    summary(model2)$r.squared,
    summary(model3)$r.squared,
    summary(model4)$r.squared,
    summary(model5)$r.squared
  ),
  N_obs = c(
    nobs(model1),
    nobs(model2),
    nobs(model3),
    nobs(model4),
    nobs(model5)
  )
)

kable(model_stats, 
      caption = "Model Comparison Summary",
      digits = 4,
      col.names = c("Model", "Unemployment Coefficient", "R²", "N"))
```

# Key Findings

## Main Results

```{r extract-coefficients}
coef_m1 <- coef(model1)["unemployment_rate"]
coef_m2 <- coef(model2)["unemployment_rate"]
coef_m4 <- coef(model4)["unemployment_rate"]
pres_effect <- coef(model2)["is_presidential"]
```

### 1. Overall Relationship (Model 1)

-   A **1 percentage point increase in unemployment** is associated with a **`r round(coef_m1, 4)` change in turnout**
-   This represents a **weak negative relationship**
-   R² = `r round(summary(model1)$r.squared, 4)` (explains very little variation)

### 2. Controlling for Election Type (Model 2)

-   Unemployment effect: **`r round(coef_m2, 4)`** (slightly smaller)
-   **Presidential elections** increase turnout by **`r round(pres_effect, 3)`** (16.3 percentage points)
-   R² = `r round(summary(model2)$r.squared, 4)` (election type explains 27% of variation!)

### 3. Interaction with County Type (Model 3)

The effect of unemployment **differs across county types**:

```{r interaction-effects}
# Extract interaction coefficients from Model 3
int_suburban <- coef(model3)["unemployment_rate:county_typeSuburban"]
int_urban <- coef(model3)["unemployment_rate:county_typeUrban"]

cat("Unemployment effect by county type:\n")
cat("  - Rural (baseline):", round(coef(model3)["unemployment_rate"], 4), "\n")
cat("  - Suburban (baseline + interaction):", 
    round(coef(model3)["unemployment_rate"] + int_suburban, 4), "\n")
cat("  - Urban (baseline + interaction):", 
    round(coef(model3)["unemployment_rate"] + int_urban, 4), "\n")
```

### 4. Panel Fixed Effects (Model 4)

-   **Within-county estimate:** `r round(coef_m4, 4)`
-   Controls for time-invariant county characteristics
-   Controls for national trends (year fixed effects)
-   **Most reliable estimate** for causal interpretation

### 5. Practical Significance

```{r practical-significance}
unemployment_change <- 2  # 2 percentage point increase
turnout_change <- coef_m4 * unemployment_change
baseline_turnout <- 0.50
new_turnout <- baseline_turnout + turnout_change
```

**Scenario:** A 2 percentage point increase in unemployment (e.g., from 5% to 7%, as during a recession)

-   Associated turnout change: **`r round(turnout_change, 4)`**
-   For a county with 50% baseline turnout:
    -   New predicted turnout: **`r round(new_turnout, 4)`** (`r round(new_turnout*100, 2)`%)

## Statistical Tests

```{r statistical-tests}
# T-test: Difference in turnout by election type
t_test_result <- t.test(VOTER_TURNOUT_PCT ~ is_presidential, data = df)

cat("T-test: Presidential vs Midterm Turnout\n")
cat("  Presidential mean:", round(t_test_result$estimate[2], 4), "\n")
cat("  Midterm mean:", round(t_test_result$estimate[1], 4), "\n")
cat("  Difference:", round(diff(t_test_result$estimate), 4), "\n")
cat("  t-statistic:", round(t_test_result$statistic, 2), "\n")
cat("  p-value:", format.pval(t_test_result$p.value, digits = 3), "\n")

# Correlation by election type
cat("\nCorrelation: Unemployment vs Turnout\n")
for (etype in c("Presidential", "Midterm")) {
  subset_df <- df %>% filter(election_type == etype)
  cor_test <- cor.test(subset_df$unemployment_rate, subset_df$VOTER_TURNOUT_PCT)
  cat("  ", etype, ": r =", round(cor_test$estimate, 4), 
      ", p =", format.pval(cor_test$p.value, digits = 3), "\n")
}
```

# Conclusions

## Research Question Answer

**How does local economic downturn relate to voter turnout?**

1.  **Weak Negative Effect:** Higher unemployment is associated with slightly lower turnout, but the effect is small

2.  **Election Type Dominates:** Presidential vs midterm elections matter 40× more than unemployment for explaining turnout

3.  **Heterogeneous Effects:** The unemployment-turnout relationship differs across urban, suburban, and rural counties

4.  **Within-County Evidence:** Fixed effects models show that when a county experiences higher unemployment, turnout decreases slightly, controlling for other factors

5.  **Practical Impact:** A recession-level increase in unemployment (2%) is associated with only a 0.4-0.8 percentage point decrease in turnout

## Limitations

-   **County-level aggregation:** Individual-level data would be preferable
-   **Omitted variables:** Cannot control for campaign spending, candidate quality, etc.
-   **Causality:** Despite fixed effects, cannot definitively establish causal relationship
-   **Time period:** Analysis limited to 2004-2022

## Policy Implications

-   Economic conditions have **modest effects** on political participation
-   Electoral reforms should focus on **structural barriers** rather than assuming economic improvements alone will boost turnout
-   **Context matters:** Urban and rural counties may respond differently to economic stress

# Appendix

## Save Results

```{r save-results}
# Create output directory
dir.create("../data/output", showWarnings = FALSE, recursive = TRUE)

# Save model objects
save(model1, model2, model3, model4, model5, 
     file = "../data/output/regression_models.RData")

# Save processed data
write_csv(df, "../data/processed/analysis_data_r.csv")

# Save regression table (LaTeX format)
stargazer(model1, model2, model3, model4, model5,
          type = "latex",
          title = "Effect of Unemployment on Voter Turnout",
          column.labels = c("OLS", "+ Election", "+ Interaction", "FE", "FE + Int"),
          dep.var.labels = "Voter Turnout Rate",
          out = "../data/output/regression_table.tex")

cat("✓ Results saved:\n")
cat("  - Model objects: data/output/regression_models.RData\n")
cat("  - Processed data: data/processed/analysis_data_r.csv\n")
cat("  - LaTeX table: data/output/regression_table.tex\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
