---
title: "Economic Downturn and Incumbent Support"
subtitle: "Research Question 2: Does local economic downturn affect support for the incumbent presidential party?"
author: "Statistical Analysis - STA 702"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    self-contained: true
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# Introduction

This analysis addresses Research Question 2:

> **Does local economic downturn affect the probability that a county supports the incumbent presidential party?**

We use the same county-year panel dataset as in Research Question 1, combining county-level unemployment rates with U.S. presidential election results from 2004–2020.

**Key Ideas:**

-   Rather than using unemployment *levels*, we focus on **unemployment relative to a county’s own long-run average** (local economic stress).
-   We study whether counties experiencing worse-than-usual unemployment are **more or less likely to vote for the incumbent presidential party**.
-   We use **fixed effects models** to compare counties to themselves over time, controlling for unobserved, time-invariant county characteristics and common national shocks.

**Key Variables (RQ2):**

-   **Outcome:** `incumbent_win` — indicator = 1 if the county votes for the incumbent presidential party, 0 otherwise\
-   **Predictor:** `unemp_change` — county unemployment rate minus the county’s historical mean unemployment

**Controls / Structure:**

-   County fixed effects (`FIPS`)\
-   Year fixed effects (`YEAR`)\
-   Optional: county type (Urban/Suburban/Rural) for heterogeneous effects

**The Core Model**

logit(Pr\[incumbent_winc,y​=1\])=β⋅unemployment_ratec,y​+αc​+γy​+ϵc,y​

Where:

# Data Preparation

```{r load-packages}
# Load required packages
library(tidyverse)
library(lfe)           # For fixed effects regression (linear probability model)
library(stargazer)     # For regression tables
library(ggplot2)
library(knitr)
library(gridExtra)

# Set ggplot theme

theme_set(theme_minimal(base_size = 12))
```

```{r load-data}
# Load the merged dataset
df <- read_csv("../data/processed/merged_voting_unemployment_complete.csv",
               show_col_types = FALSE)

# Display data structure
cat("Dataset Overview:\n")
cat("Observations:", nrow(df), "\n")
cat("Counties:", length(unique(df$FIPS)), "\n")
cat("Years:", min(df$YEAR), "-", max(df$YEAR), "\n")
cat("Variables:", ncol(df), "\n")
```

## Variable Creation

```{r}
names(df)
```

```{r}
#Create incumbent-party lookup table
incumbent_table <- tibble(
  YEAR = c(2000, 2004, 2008, 2012, 2016, 2020),
  incumbent_party = c("DEM", "REP", "REP", "DEM", "DEM", "REP")
)

#Merge with df
df <- df %>%
  left_join(incumbent_table, by = "YEAR")

#Simplify winner in each County
df <- df %>%
  mutate(
    winner_party = case_when(
      PRES_DEM_VOTES > PRES_REP_VOTES ~ "DEM",
      PRES_REP_VOTES > PRES_DEM_VOTES ~ "REP",
      TRUE ~ NA_character_  # in case of ties or missing
    )
  )


```

```{r}
#Creation of the variable incumbent_win

df <- df %>%
  mutate(
    incumbent_win = if_else(winner_party == incumbent_party, 1, 0)
  )


```

```{r}
#Checking results
table(df$incumbent_win, useNA = "ifany")
head(df %>% select(YEAR, PRES_DEM_VOTES, PRES_REP_VOTES, winner_party, incumbent_party, incumbent_win))

```

```{r}
#Keep only presidential election years

df_pres <- df %>% filter(YEAR %in% c(2000, 2004, 2008, 2012, 2016, 2020))


```

```{r}
library(dplyr)

df_pres <- df %>%
  filter(
    YEAR %in% c(2000, 2004, 2008, 2012, 2016, 2020),
    !is.na(incumbent_win),
    !is.na(unemployment_rate)
  )

nrow(df_pres)
head(df_pres %>% select(YEAR, FIPS, incumbent_win, unemployment_rate))

```

# **Model Specification**

logit(Pr\[incumbent_winc,y​=1\])=β⋅unemployment_ratec,y​+αc​+γy​+ϵc,y​.

We estimate a logistic regression model to assess whether local economic conditions, measured through the county–year unemployment rate, influence the probability that a county votes for the incumbent presidential party. The outcome variable is *incumbent_win* (1 if the incumbent party carried the county, 0 otherwise). To isolate the effect of unemployment from stable county-level differences (e.g., political lean, demographics) and from national election shocks, we include county fixed effects and year fixed effects. This aligns with standard practice in political economy research using panel data.

```{r}
install.packages("fixest")  
library(fixest)

```

```{r}
#Fixed Logistic Model

model_rq2 <- feglm(
  incumbent_win ~ unemployment_rate | FIPS + YEAR,
  family = binomial(link = "logit"),
  data = df_pres
)

```

```{r}
#Model Summary

summary(model_rq2)

```

The coefficient on **unemployment_rate** is **negative, statistically significant, and substantively meaningful** (β = –0.096, p \< 0.001). This indicates that *higher unemployment in a county is associated with a lower probability that the incumbent party wins that county*. In other words, when economic conditions worsen locally, incumbents tend to perform worse. Because the model is logistic, this coefficient reflects the log-odds of an incumbent win; in practical terms, a one-percentage-point increase in the unemployment rate reduces the odds of an incumbent victory by approximately **9–10%**, holding constant both county characteristics and national election dynamics.

This finding aligns strongly with classical theories of **retrospective economic voting**, which suggest that voters tend to reward incumbents during periods of economic prosperity and punish them during downturns. Importantly, our model’s fixed-effects structure ensures that this pattern is not driven by long-standing partisan differences between counties (which are absorbed by county fixed effects) or by national phenomena such as popular incumbent presidents, economic crises, or unusually high-turnout years (absorbed by year fixed effects). What remains is the *localized* economic signal: whether conditions in that specific county are better or worse than usual.

The magnitude of the effect is modest but meaningful. Presidential elections involve complex, multi-dimensional decision processes, and no single variable can fully determine outcomes. However, the fact that unemployment consistently and significantly predicts incumbent losses suggests that local economic pain plays a nontrivial role in shaping electoral behavior. Even small shifts in county-level probabilities can aggregate into impactful changes across swing states or competitive regions.

Overall, these results provide clear evidence that **economic downturns at the county level reduce the electoral fortunes of the incumbent presidential party**. This supports both retrospective voting theory and broader political behavior frameworks that argue economic performance is a central driver of voter accountability and political punishment.

# **Plots and Visualizations**

```{r}
ggplot(df_pres, aes(x = unemployment_rate, y = incumbent_win)) +
  geom_jitter(height = 0.05, width = 0, alpha = 0.25, color = "#2C3E50") +
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"),
              color = "#E74C3C", 
              se = TRUE) +
  labs(
    title = "Incumbent Victory vs. County Unemployment Rate",
    x = "Unemployment Rate (%)",
    y = "Probability of Incumbent Win"
  ) +
  theme_minimal(base_size = 13)

```
