---
title: "Economic Downturn and Incumbent Support"
subtitle: "Research Question 2: Does local economic downturn affect support for the incumbent presidential party?"
author: "Statistical Analysis - STA 702"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    self-contained: true
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# Introduction

This analysis addresses Research Question 2:

> **Does local economic downturn affect the probability that a county supports the incumbent presidential party?**

We use the same county-year panel dataset as in Research Question 1, combining county-level unemployment rates with U.S. presidential election results from 2004–2020.

**Key Ideas:**

-   Rather than using unemployment *levels*, we focus on **unemployment relative to a county’s own long-run average** (local economic stress).
-   We study whether counties experiencing worse-than-usual unemployment are **more or less likely to vote for the incumbent presidential party**.
-   We use **fixed effects models** to compare counties to themselves over time, controlling for unobserved, time-invariant county characteristics and common national shocks.

**Key Variables (RQ2):**

-   **Outcome:** `incumbent_win` — indicator = 1 if the county votes for the incumbent presidential party, 0 otherwise\
-   **Predictor:** `unemp_change` — county unemployment rate minus the county’s historical mean unemployment

**Controls / Structure:**

-   County fixed effects (`FIPS`)\
-   Year fixed effects (`YEAR`)\
-   Optional: county type (Urban/Suburban/Rural) for heterogeneous effects

**The Core Model**

logit(Pr\[incumbent_winc,y​=1\])=β⋅unemployment_ratec,y​+αc​+γy​+ϵc,y​

Where:

# Data Preparation

```{r load-packages}
# Load required packages
library(tidyverse)
library(lfe)           # For fixed effects regression (linear probability model)
library(stargazer)     # For regression tables
library(ggplot2)
library(knitr)
library(gridExtra)

# Set ggplot theme

theme_set(theme_minimal(base_size = 12))
```

```{r load-data}
# Load the merged dataset
df <- read_csv("../data/processed/merged_voting_unemployment_complete.csv",
               show_col_types = FALSE)

# Display data structure
cat("Dataset Overview:\n")
cat("Observations:", nrow(df), "\n")
cat("Counties:", length(unique(df$FIPS)), "\n")
cat("Years:", min(df$YEAR), "-", max(df$YEAR), "\n")
cat("Variables:", ncol(df), "\n")
```

## Variable Creation

```{r}
names(df)
```

```{r}
#Create incumbent-party lookup table
incumbent_table <- tibble(
  YEAR = c(2000, 2004, 2008, 2012, 2016, 2020),
  incumbent_party = c("DEM", "REP", "REP", "DEM", "DEM", "REP")
)

#Merge with df
df <- df %>%
  left_join(incumbent_table, by = "YEAR")

#Simplify winner in each County
df <- df %>%
  mutate(
    winner_party = case_when(
      PRES_DEM_VOTES > PRES_REP_VOTES ~ "DEM",
      PRES_REP_VOTES > PRES_DEM_VOTES ~ "REP",
      TRUE ~ NA_character_  # in case of ties or missing
    )
  )


```

```{r}
#Creation of the variable incumbent_win

df <- df %>%
  mutate(
    incumbent_win = if_else(winner_party == incumbent_party, 1, 0)
  )


```

```{r}
#Checking results
table(df$incumbent_win, useNA = "ifany")
head(df %>% select(YEAR, PRES_DEM_VOTES, PRES_REP_VOTES, winner_party, incumbent_party, incumbent_win))

```

```{r}
#Keep only presidential election years

df_pres <- df %>% filter(YEAR %in% c(2000, 2004, 2008, 2012, 2016, 2020))


```
