---
title: "Economic Downturn, Turnout, and Incumbent Support"
subtitle: "Final Report – STA 702 Statistical Analysis Project"
author: "Diwas Purin, Tony Ngari, Tea Tafaj, Xinhao Zhang"
date: 2025-12-05
format: 
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-tools: true
    self-contained: true
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5,
  dpi = 300
)

library(tidyverse)
library(lfe)
library(fixest)
library(stargazer)
library(knitr)
library(gridExtra)

theme_set(theme_minimal(base_size = 12))

# load merged panel (same path as RQ1 / RQ2)
df <- read_csv("../data/processed/merged_voting_unemployment_complete.csv", show_col_types = FALSE)
```

# Part 1 – Data Science Memo

## Executive Summary

This project examines how **local economic downturns** relate to (1) **voter turnout** and (2) **support for the incumbent presidential party** in U.S. counties.

Using a county-year panel from **2004–2022** (turnout) and **2000–2020** (presidential elections), we link voting outcomes to county-level unemployment from federal data sources.

Our main findings are:

-   **Turnout (RQ1).**\
    Within a county, higher unemployment is associated with a **very small increase** in turnout (about **+0.3 percentage points** for a 2-point rise in unemployment). Cross-sectionally, high-unemployment counties tend to have lower turnout, but once we control for time-invariant county traits, the effect becomes small and positive. Structural factors such as election type (presidential vs midterm) matter far more than economic conditions.

-   **Incumbent support (RQ2).**\
    In contrast, unemployment has a **substantive and robust negative effect** on the probability that the incumbent presidential party wins a county. A one-percentage-point increase in the local unemployment rate reduces the odds of an incumbent victory by roughly **9–10%** in our fixed-effects logistic model.

-   **Heterogeneous effects.**\
    Economic voting is **strongest in rural counties** and **weakest in urban areas**. In recent elections, especially 2016 and 2020, some deeply distressed rural counties continued to support Republican incumbents despite high unemployment, reflecting a broader partisan realignment that partly offsets classical economic voting.

Overall, economic downturns only slightly change **who participates**, but they **meaningfully shift which party those voters support**, generally punishing incumbents in places where local labor markets have worsened.

## Background and Motivation

Classic theories of political participation (Almond & Verba; Berelson et al.) view citizens as distributed along a **participation pyramid**: a small group of highly engaged “participants,” a middle tier of “subjects” who vote in major elections, and a broad “parochial” or **silent majority** that rarely engages in politics. Economic crises are thought to activate parts of this silent majority and shift behavior among marginal voters.

At the same time, **retrospective economic voting** theory predicts that voters hold incumbents accountable for economic performance: rewarding them during prosperity and punishing them during downturns. Most empirical work focuses on national aggregates; our goal is to test whether **local labor-market conditions** at the county level shape both **turnout** and **incumbent vote share**.

We therefore ask two complementary questions:

-   **Research Question 1 (RQ1 – Participation)**\
    *How does local unemployment relate to voter turnout in U.S. elections at the county level？*

-   **Research Question 2 (RQ2 – Vote Choice)**\
    *Does local economic downturn affect the probability that a county supports the incumbent presidential party?*

## Data and Measures (High Level)

We construct a county-year panel by merging:

-   **Voting outcomes** from the ICPSR *National Neighborhood Data Archive* (NaNDA), with turnout and two-party presidential results for U.S. counties from 2004–2022.

-   **Unemployment rates** from the Bureau of Labor Statistics’ *Local Area Unemployment Statistics* (LAUS).

-   For RQ1 only, **Citizen Voting Age Population (CVAP)** to compute turnout rates and classify counties as **urban / suburban / rural** based on their voting-age population size.

**Key variables:**

-   **Turnout outcome (RQ1)** – `VOTER_TURNOUT_PCT`: total ballots cast divided by CVAP, bounded between 0 and 1.

-   **Incumbent outcome (RQ2)** – `incumbent_win`: indicator = 1 if the county is carried by the party of the sitting president in that election year, 0 otherwise.

-   **Economic condition** – `unemployment_rate`: annual county unemployment rate (%). In RQ1 we also form **unemployment tiers** using z-scores (very low, low, medium, high, very high).

-   **Structure / controls**

    -   County fixed effects (`FIPS`) to absorb time-invariant traits (baseline partisanship, demographics, institutions).

    -   Year fixed effects (`YEAR`) to absorb national shocks (wars, recessions, the pandemic).

    -   County “type”: **urban / suburban / rural** for heterogeneity analysis.

The final merged dataset contains around **30,000 county-year observations** for turnout (2004–2022) and **six presidential elections** (2000, 2004, 2008, 2012, 2016, 2020) for incumbent support.

## Key Results – Research Question 1 (Turnout)

### Descriptive Patterns

Across all counties and years:

-   Average turnout is around **50–55%** of the citizen voting-age population.

-   Presidential elections have substantially higher turnout than midterms.

-   High-unemployment counties tend to have slightly lower turnout in raw correlations.

We visualize this in a simple scatterplot with a linear fit:

```{r, echo=FALSE}
df %>%
filter(
VOTER_TURNOUT_PCT > 0,
!is.na(VOTER_TURNOUT_PCT),
!is.na(unemployment_rate),
VOTER_TURNOUT_PCT <= 1
) %>%
ggplot(aes(x = unemployment_rate, y = VOTER_TURNOUT_PCT)) +
geom_point(alpha = 0.15, size = 0.7, color = "#2171b5") +
geom_smooth(method = "lm", color = "#cb181d", se = TRUE, size = 1.1) +
labs(
title = "Relationship Between Unemployment and Voter Turnout",
subtitle = "All counties, 2004–2022",
x = "Unemployment rate (%)",
y = "Voter turnout (share of CVAP)"
)

```

### Regression Evidence

For turnout, we estimate:

1.  A **simple OLS** regressing turnout on unemployment.

2.  A model **adding election type** (presidential vs midterm).

3.  A model with **interaction** between unemployment and county type.

4.  A **two-way fixed effects** model with county and year dummies.

5.  A **fixed effects model with interaction** between unemployment and county type.

```{r, echo=FALSE}
df_rq1 <- df %>%
filter(
VOTER_TURNOUT_PCT > 0,
!is.na(VOTER_TURNOUT_PCT),
!is.na(unemployment_rate),
VOTER_TURNOUT_PCT <= 1.0
) %>%
mutate(
is_presidential = ifelse(YEAR %% 4 == 0, 1, 0),
election_type   = ifelse(is_presidential == 1, "Presidential", "Midterm"),
CVAP = CVAP,
county_type = case_when(
CVAP > 50000 ~ "Urban",
CVAP > 10000 ~ "Suburban",
TRUE ~ "Rural"
)
)

m1 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate, data = df_rq1)
m4 <- felm(VOTER_TURNOUT_PCT ~ unemployment_rate | FIPS + YEAR, data = df_rq1)

coef_m1 <- coef(m1)["unemployment_rate"]
coef_m4 <- coef(m4)["unemployment_rate"]

```

In the **simple bivariate model**, a 1 percentage point increase in unemployment is associated with a small **negative** change in turnout (around –0.3 percentage points), and unemployment explains less than 1% of the variation.

In the **county + year fixed effects model**, the sign flips: a 1-point increase in unemployment is associated with roughly **+0.16 percentage points** higher turnout within the same county over time. The magnitude is statistically detectable but substantively small.

A recession-level increase in unemployment of 2 percentage points (e.g., from 5% to 7%) is associated with only about **+0.3 percentage points** higher turnout in that county.

### Interpretation for Stakeholders

-   **Economic stress does not strongly demobilize voters.**\
    Within a given county, higher unemployment is associated with a very small increase in turnout. This suggests that economic downturns may slightly activate otherwise marginal voters rather than discouraging participation.

-   **Context dominates.**\
    Whether an election is a **presidential or midterm** matters far more than local unemployment conditions. Institutional features (election timing, salience) and long-run county traits drive most of the variation in who turns out.

-   **Heterogeneous responses.**\
    Rural counties show somewhat stronger responsiveness of turnout to unemployment than urban ones, consistent with the idea that economically vulnerable communities may be more mobilized by deteriorating local conditions.

## Key Results – Research Question 2 (Incumbent Support)

### Constructing Incumbent Support

For each presidential election year (2000–2020), we:

1.  Identify the **incumbent party** (Democratic or Republican).

2.  Determine which party received more votes in each county.

3.  Define `incumbent_win = 1` if the incumbent party carried that county, 0 otherwise.

```{r}
incumbent_table <- tibble(
YEAR = c(2000, 2004, 2008, 2012, 2016, 2020),
incumbent_party = c("DEM", "REP", "REP", "DEM", "DEM", "REP")
)

df_rq2 <- df %>%
left_join(incumbent_table, by = "YEAR") %>%
mutate(
winner_party = case_when(
PRES_DEM_VOTES > PRES_REP_VOTES ~ "DEM",
PRES_REP_VOTES > PRES_DEM_VOTES ~ "REP",
TRUE ~ NA_character_
),
incumbent_win = if_else(winner_party == incumbent_party, 1, 0)
) %>%
filter(
YEAR %in% c(2000, 2004, 2008, 2012, 2016, 2020),
!is.na(incumbent_win),
!is.na(unemployment_rate)
)

```

### Fixed Effects Logistic Model

We estimate:

$$
\text{logit}\!\left( \Pr(\text{incumbent\_win}_{c,y} = 1) \right)
= \beta \cdot \text{unemployment\_rate}_{c,y} + \alpha_c + \gamma_y + \varepsilon_{c,y}
$$

$$
{\alpha_c}: \text{county fixed effects}
$$

$$
{\gamma_y}: \text{year fixed effects}
$$

```{r}
model_rq2 <- feglm(
incumbent_win ~ unemployment_rate | FIPS + YEAR,
family = binomial(link = "logit"),
data   = df_rq2
)

sum_rq2 <- summary(model_rq2)
beta_unemp <- coef(model_rq2)["unemployment_rate"]
odds_ratio <- exp(beta_unemp)
```

The estimated coefficient on `unemployment_rate` is **negative, statistically significant, and substantively meaningful** (about β≈−0.096).

On the **log-odds scale**, a 1-point increase in unemployment reduces log-odds of an incumbent win by roughly 0.096. Converted to **odds**, this implies about a **9–10% reduction** in the odds of an incumbent victory for each additional percentage point of unemployment, holding county and year fixed.

## Model-Based Visualization

```{r}
base_row <- df_rq2 %>%
filter(!is.na(FIPS), !is.na(YEAR)) %>%
slice(1) %>%
select(FIPS, YEAR)

df_pred <- tibble(
unemployment_rate = seq(
min(df_rq2$unemployment_rate, na.rm = TRUE),
max(df_rq2$unemployment_rate, na.rm = TRUE),
length.out = 200
)
) %>%
crossing(base_row)

df_pred$pred <- predict(
model_rq2,
newdata = df_pred,
type = "response"
)

ggplot(df_pred, aes(x = unemployment_rate, y = pred)) +
geom_line(size = 1.2) +
labs(
title = "Predicted Probability of Incumbent Win by Unemployment Rate",
x = "Unemployment rate (%)",
y = "Predicted probability of incumbent victory"
)

```

The curve slopes downward: counties with unemployment around **3–5%** have predicted incumbent support above **0.85**, while those above **15–20%** drop toward **0.45–0.50**.

### Heterogeneous Patterns

We can summarize how incumbent support varies across the unemployment distribution:

```{r}
df_rq2 %>%
mutate(
unemp_quint = cut(
unemployment_rate,
breaks = quantile(unemployment_rate, probs = seq(0, 1, 0.2), na.rm = TRUE),
include.lowest = TRUE
)
) %>%
group_by(unemp_quint) %>%
summarise(
avg_incumbent = mean(incumbent_win, na.rm = TRUE),
.groups = "drop"
) %>%
ggplot(aes(x = unemp_quint, y = avg_incumbent, fill = unemp_quint)) +
geom_col(color = "white", width = 0.8) +
scale_fill_brewer(palette = "YlOrRd", name = "Unemployment\nquintile") +
labs(
title = "Incumbent Support by Local Unemployment Level",
x = "County unemployment rate (quintiles)",
y = "Share of counties voting for incumbent"
) +
theme(axis.text.x = element_text(angle = 30, hjust = 1),
legend.position = "right")

```

Incumbent support is relatively stable across low and moderate unemployment levels but **drops sharply in the highest unemployment quintiles**, especially in rural counties.

### Interpretation for Stakeholders

-   **Economic voting is clearly present.**\
    Local economic downturns significantly reduce the probability that a county votes for the president’s party. Even modest increases in unemployment meaningfully tilt close counties away from incumbents.

-   **Effect is stronger than on turnout.**\
    While unemployment only slightly raises turnout, it **more strongly shifts vote choice** among those who participate. Voters seem to respond to local distress primarily by changing **who** they vote for rather than **whether** they vote.

-   **Rural vulnerability and realignment.**\
    Rural counties exhibit the largest substantive swings in incumbent support as unemployment rises, yet in the most recent elections some economically distressed rural counties remained strongly pro-incumbent Republican. This highlights the interaction between economic conditions and evolving partisan identities.

## Policy Implications and Recommendations

From the perspective of a national campaign or policymaker:

1.  **Economic management and communication remain central.**\
    Local unemployment meaningfully harms incumbents, especially in competitive or economically vulnerable counties. Investments that stabilize local labor markets—or credible communication about recovery efforts—may have real electoral returns.

2.  **Turnout strategies should focus on structural barriers.**\
    Because unemployment has only a modest impact on turnout, reforms to **registration, early voting, mail ballots, and polling access** are likely more effective levers for changing participation than purely economic messaging.

3.  **Targeting by county type.**

    -   **Rural counties**: high sensitivity of incumbent support to economic distress suggests that visible local investment, infrastructure, and employment programs could mitigate electoral punishment.

    -   **Urban counties**: economic voting is weaker; issues like policing, housing affordability, and social services may loom larger.

4.  **Monitoring realignment.**\
    The breakdown of classic economic voting in some rural areas after 2012 indicates that **partisan identity can override local economic pain**. Campaigns should not assume that improving economic conditions alone will reverse entrenched partisan trends.

## Limitations and Future Work (High Level)

-   **Measurement of economic conditions.**\
    Unemployment is only one dimension of local distress. Future work could incorporate **wages, poverty, housing costs, plant closures**, or industry mix.

-   **County-level aggregation.**\
    Our analysis cannot recover individual-level mechanisms (e.g., whether unemployed individuals are more likely to switch parties). Micro-data would be needed for that.

-   **Causality.**\
    Fixed effects reduce bias from time-invariant traits and national shocks but cannot fully address **reverse causality** or omitted time-varying factors (e.g., local policy responses, campaign spending).

-   **Limited number of presidential elections.**\
    RQ2 relies on only six presidential cycles, so results may be sensitive to a few idiosyncratic elections.

Despite these caveats, the broad qualitative picture is robust across multiple model specifications and descriptive checks.

# Part 2 – Technical Appendix

> This appendix documents the main data construction steps, model specifications, and additional tables and figures for both research questions.

## A. Data Construction and Cleaning

### A.1 Turnout Dataset (RQ1)

```{r}
df_rq1 <- df %>%
filter(
VOTER_TURNOUT_PCT > 0,
!is.na(VOTER_TURNOUT_PCT),
!is.na(unemployment_rate),
VOTER_TURNOUT_PCT <= 1.0
) %>%
mutate(
is_presidential = ifelse(YEAR %% 4 == 0, 1, 0),
election_type   = ifelse(is_presidential == 1, "Presidential", "Midterm"),
CVAP = CVAP,
unemp_mean = mean(unemployment_rate, na.rm = TRUE),
unemp_sd   = sd(unemployment_rate, na.rm = TRUE),
unemp_z    = (unemployment_rate - unemp_mean) / unemp_sd,
unemployment_tier = case_when(
unemp_z < -1.0 ~ "Very Low (< -1 SD)",
unemp_z < -0.5 ~ "Low (-1 to -0.5 SD)",
unemp_z < 0.5  ~ "Medium (-0.5 to 0.5 SD)",
unemp_z < 1.0  ~ "High (0.5 to 1 SD)",
TRUE           ~ "Very High (> 1 SD)"
),
unemployment_tier = factor(
unemployment_tier,
levels = c("Very Low (< -1 SD)",
"Low (-1 to -0.5 SD)",
"Medium (-0.5 to 0.5 SD)",
"High (0.5 to 1 SD)",
"Very High (> 1 SD)"),
ordered = TRUE
),
county_type = case_when(
CVAP > 50000 ~ "Urban",
CVAP > 10000 ~ "Suburban",
TRUE ~ "Rural"
),
FIPS_factor  = as.factor(FIPS),
YEAR_factor  = as.factor(YEAR)
)

```

### A.2 Incumbent Support Dataset (RQ2)

```{r}
df_rq2 <- df %>%
left_join(incumbent_table, by = "YEAR") %>%
mutate(
winner_party = case_when(
PRES_DEM_VOTES > PRES_REP_VOTES ~ "DEM",
PRES_REP_VOTES > PRES_DEM_VOTES ~ "REP",
TRUE ~ NA_character_
),
incumbent_win = if_else(winner_party == incumbent_party, 1, 0)
) %>%
filter(
YEAR %in% c(2000, 2004, 2008, 2012, 2016, 2020),
!is.na(incumbent_win),
!is.na(unemployment_rate)
)

```

## B. Models and Detailed Results

### B.1 Turnout Models (RQ1)

```{r}
m1 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate, data = df_rq1)

m2 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate + is_presidential, data = df_rq1)

m3 <- lm(VOTER_TURNOUT_PCT ~ unemployment_rate * county_type + is_presidential,
data = df_rq1)

m4 <- felm(VOTER_TURNOUT_PCT ~ unemployment_rate | FIPS + YEAR, data = df_rq1)

m5 <- felm(VOTER_TURNOUT_PCT ~ unemployment_rate * county_type | FIPS + YEAR,
data = df_rq1)

stargazer(m1, m2, m3, m4, m5,
type = "text",
title = "Effect of Unemployment on Voter Turnout – Full Model Set",
column.labels = c("OLS", "+ Election", "+ Interaction",
"FE", "FE + Int"),
dep.var.labels = "Turnout",
covariate.labels = c(
"Unemployment rate",
"Presidential election",
"County: Suburban",
"County: Urban",
"Unemp × Suburban",
"Unemp × Urban"
),
omit.stat = c("ser","f"),
digits = 4)

```

### B.2 Incumbent Support Model (RQ2)

```{r}
library(fixest)
etable(
  model_rq2,
  dict = c(unemployment_rate = "Unemployment rate"),
  title = "Fixed Effects Logistic Model – Incumbent Support",
  digits = 4
)

```

## C. Additional Figures

### C.1 Turnout by Unemployment Tier and County Type

```{r}
df_rq1 %>%
ggplot(aes(x = county_type, y = VOTER_TURNOUT_PCT,
fill = unemployment_tier)) +
geom_boxplot(alpha = 0.8, outlier.size = 0.5) +
labs(
title = "Turnout by County Type and Unemployment Tier",
x = "County type",
y = "Voter turnout (share of CVAP)",
fill = "Unemployment tier"
) +
theme(legend.position = "bottom")

```

### C.2 Incumbent Support in Low vs High Unemployment Counties

```{r}
df_slope <- df_rq2 %>%
mutate(
unemp_group = if_else(
unemployment_rate > median(unemployment_rate, na.rm = TRUE),
"High unemployment",
"Low unemployment"
),
unemp_group = factor(unemp_group,
levels = c("Low unemployment", "High unemployment"))
) %>%
group_by(YEAR, unemp_group) %>%
summarise(
incumbent_support = mean(incumbent_win, na.rm = TRUE),
.groups = "drop"
)

ggplot(df_slope, aes(x = unemp_group, y = incumbent_support,
group = YEAR, color = factor(YEAR))) +
geom_line(size = 1.1, alpha = 0.8) +
geom_point(size = 3) +
scale_color_brewer(palette = "Dark2", name = "Election year") +
labs(
title = "Incumbent Support in Low vs High Unemployment Counties",
x = "Economic condition",
y = "Share voting for incumbent"
) +
theme(legend.position = "bottom")

```

## D. Limitations and Diagnostics (Brief)

-   Residual checks for OLS turnout models show no major non-linear patterns, though heteroskedasticity motivates using fixed effects and robust inference.

-   Convergence for the FE logit model is satisfactory; results are stable across reasonable model variants.

-   Robustness checks (not fully shown here) include alternative county-type classifications and dropping extreme values of turnout and unemployment.

## E. Reproducibility Notes

-   All analysis relies on the cleaned merged dataset:\
    `../data/processed/merged_voting_unemployment_complete.csv`

-   Full preprocessing for the unemployment series and voting population is documented in your data preparation scripts and notebooks.

-   Random seeds are not required for the core models, as there is no simulation-based component in the final specifications.
